---
title: "GSE124073_defScl"
author: "José Manuel García Pérez"
date: "2026-06-20"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

# Structures

## Experimental Design / Metadata

```{r}
# Load file
metadata = read.table(file = "data/Scl_GSE124073_def/Metadata_GSE124073_Utrecht.txt", 
                      header = T, sep = "\t")

# Available information
colnames(metadata)
rownames(metadata) = metadata$Sample

# Select relevant information
metadata = metadata[,c("Sample", "Run", "Condition", "Subtype")]

# Adapt format
metadata$Condition  = factor(metadata$Condition)
metadata$Subtype    = factor(metadata$Subtype, levels = c("hc", "non_cut", "limited", "diffuse"))

# Info summary
summary(metadata)
```

## From transcript to gene level

```{r}
library(rtracklayer)
library(tximport)

  # # Get tx2gene df from the original annotation file used
  # gtf_df = as.data.frame(import("data/Homo_sapiens.GRCh38.115.chr.gtf.gz"))
  # tx2gene = gtf_df[gtf_df$type == "transcript", c("transcript_id", "gene_id")]
  # tx2gene = tx2gene[!duplicated(tx2gene$transcript_id), ]
  # rownames(tx2gene) = NULL
  # saveRDS(tx2gene, "data/tx2gene_Ensembl115.rds")

# Load the tx2gene df
tx2gene = readRDS("data/tx2gene_Ensembl115.rds")
head(tx2gene)

# Path to the quantification files
quant_files = file.path("data/Scl_GSE124073_def/counts",
                        paste0(metadata$Sample, "_quant.sf"))

# Import and collapse to gene level
txi = tximport(quant_files, type = "salmon",
               tx2gene = tx2gene, ignoreTxVersion = TRUE)

# Check the final number of genes and samples
dim(txi$counts)
```

## The DESeqDataSet

```{r}
# Create the first basic DESeqDataSet
library(DESeq2)
dds = DESeqDataSetFromTximport(txi     = txi,
                               colData = metadata,
                               design  = ~ Condition)

# Ensure the reference level
dds$Condition = relevel(x = dds$Condition, ref = "hc")
```

# Exploratory Analysis

## Pre-filtering low counts / Gene-level QC

```{r}
# Smallest group?
table(dds$Condition, dds$Subtype)

# Keep genes with at least 10 counts for the smallest group size
keep = rowSums(counts(dds) >= 10) >= 7
dds = dds[keep,]

# How many samples are removed?
table(keep)
```

## Sample-level QC

### Data Transformation

```{r}
library("vsn")
library("ggplot2")

# "basic" log2(n+1) transformation
ntd_dds = normTransform(dds)

# Variance Stabilizing Transformation (VST)
vst_dds = vst(dds, blind=TRUE)

# Compare effect on the variance
var_raw = meanSdPlot(counts(dds), plot = FALSE)$gg + ggtitle("Raw Counts")
var_raw

var_ntd = meanSdPlot(assay(ntd_dds), plot = FALSE)$gg + ggtitle("log2(n+1)")
var_ntd

var_vst = meanSdPlot(assay(vst_dds), plot = FALSE)$gg + ggtitle("VST")
var_vst

# Final improvement in distribution check
boxplot(counts(dds), outline=T, 
        main ="Raw Data", xaxt="n",
        xlab = paste0("Samples, n=", nrow(metadata)),
        ylab = "Raw Counts",
        col  = ifelse(metadata$Condition == "hc", "skyblue", "salmon"),
        ylim = c(min(counts(dds))-1, max(counts(dds))+3))
legend(x = "topright", legend = c("Control", "Target"), 
       fill = c("skyblue", "salmon"), bty = "n", cex = 0.8)


boxplot(assay(vst_dds), outline=T, 
        main ="VST", xaxt="n",
        xlab = paste0("Samples, n=", nrow(metadata)),
        ylab = "Transformed Counts",
        col  = ifelse(metadata$Condition == "hc", "skyblue", "salmon"),
        ylim = c(min(assay(vst_dds))-1, max(assay(vst_dds))+3))
legend(x = "topright", legend = c("Control", "Target"), 
       fill = c("skyblue", "salmon"), bty = "n", cex = 0.8)
```

### Principal Components Analysis (PCA)

```{r}
# Condition
plotPCA(vst_dds,
        intgroup = c("Condition"),
        ntop = 500,
        pcsToUse = 1:2) + 
  stat_ellipse(aes(fill = group), geom = "polygon", alpha = 0) +
  scale_color_manual(values = c("skyblue","salmon"))

# Deep exploration - potential outliers
library(ggrepel)

for (i in c(500,2000,10000)){
  p = plotPCA(vst_dds,
              intgroup = c("Subtype"),
              ntop = i,
              pcsToUse = 1:2) + 
      geom_text_repel(aes(label = name), size = 3) +
      stat_ellipse(aes(fill = group), geom = "polygon", alpha = 0) +
      scale_color_manual(values = c("skyblue","palegreen3","plum","coral2")) +
      ggtitle(paste0("Top-",i))
  print(p)
}  

for (i in c(500,2000,10000)){
  p = plotPCA(vst_dds,
              intgroup = c("Subtype"),
              ntop = i,
              pcsToUse = c(1,3)) + 
      geom_text_repel(aes(label = name), size = 3) +
      stat_ellipse(aes(fill = group), geom = "polygon", alpha = 0) +
      scale_color_manual(values = c("skyblue","palegreen3","plum","coral2")) +
      ggtitle(paste0("Top-",i))
  print(p)
}  

for (i in c(500,2000,10000)){
  p = plotPCA(vst_dds,
              intgroup = c("Subtype"),
              ntop = i,
              pcsToUse = c(2,3)) + 
      geom_text_repel(aes(label = name), size = 3) +
      stat_ellipse(aes(fill = group), geom = "polygon", alpha = 0) +
      scale_color_manual(values = c("skyblue","palegreen3","plum","coral2")) +
      ggtitle(paste0("Top-",i))
  print(p)
}  

# Specific comparisons

for (i in c(500,2000,10000)){
  p = plotPCA(vst_dds[,vst_dds$Subtype %in% c("hc","non_cut")],
              intgroup = c("Subtype"),
              ntop = i,
              pcsToUse = 1:2) + 
      geom_text_repel(aes(label = name), size = 3) +
      stat_ellipse(aes(fill = group), geom = "polygon", alpha = 0) +
      scale_color_manual(values = c("skyblue","palegreen3")) +
      ggtitle(paste0("Top-",i))
  print(p)
} 

for (i in c(500,2000,10000)){
  p = plotPCA(vst_dds[,vst_dds$Subtype %in% c("hc","limited")],
              intgroup = c("Subtype"),
              ntop = i,
              pcsToUse = 1:2) + 
      geom_text_repel(aes(label = name), size = 3) +
      stat_ellipse(aes(fill = group), geom = "polygon", alpha = 0) +
      scale_color_manual(values = c("skyblue","plum")) +
      ggtitle(paste0("Top-",i))
  print(p)
}

for (i in c(500,2000,10000)){
  p = plotPCA(vst_dds[,vst_dds$Subtype %in% c("hc","diffuse")],
              intgroup = c("Subtype"),
              ntop = i,
              pcsToUse = 1:2) + 
      geom_text_repel(aes(label = name), size = 3) +
      stat_ellipse(aes(fill = group), geom = "polygon", alpha = 0) +
      scale_color_manual(values = c("skyblue","coral2")) +
      ggtitle(paste0("Top-",i))
  print(p)
}
```

### Sample clustering

```{r}
library(pheatmap)
library(RColorBrewer)

# Define groups and colors
df_anno = as.data.frame(metadata[, c("Condition", "Subtype")])

ann_colors = list(
  Condition  = c(hc = "skyblue", scl = "salmon"),
  Subtype    = c(hc = "skyblue", non_cut = "palegreen3", 
                 limited = "plum", diffuse ="coral2")
)

# Distance-based heatmap
dist_matrix = as.matrix(dist(t(assay(vst_dds))))

pheatmap(dist_matrix,
         col = colorRampPalette(rev(brewer.pal(9, "Reds")))(255),
         clustering_method = "ward.D2",
         clustering_distance_rows = as.dist(dist_matrix),
         clustering_distance_cols = as.dist(dist_matrix),
         treeheight_col = 0,
         annotation_row = df_anno, annotation_col = df_anno,
         annotation_colors = ann_colors,
         show_colnames = FALSE, show_rownames = FALSE,
         annotation_names_row = FALSE,
         annotation_names_col = FALSE,
         main = "Sample-to-Sample Distance (VST)",
         border_color = NA)

# Correlation-based heatmap
cor_matrix = cor(assay(vst_dds), method = "pearson")

# Create fixed color fade to compare after QC
n_breaks = 255
breaks_fixed = seq(min(cor_matrix), max(cor_matrix), length.out = n_breaks+1)
colors_fixed = colorRampPalette(brewer.pal(9, "Reds"))(n_breaks)

pheatmap(cor_matrix,
         breaks = breaks_fixed,
         col = colors_fixed,
         clustering_method = "ward.D2",
         clustering_distance_rows = as.dist(1 - cor_matrix),
         clustering_distance_cols = as.dist(1 - cor_matrix),
         treeheight_col = 0,
         annotation_row = df_anno, annotation_col = df_anno,
         annotation_colors = ann_colors,
         show_colnames = FALSE, show_rownames = FALSE,
         annotation_names_row = FALSE,
         annotation_names_col = FALSE,
         main = "Sample-to-Sample Correlation (VST)",
         border_color = NA)

# Find the correlation outlier(s)
mean_cor = colMeans(cor_matrix)
sort(mean_cor)[1:10]


#### Subsets

## Non cut

target_samples = colData(vst_dds)$Subtype %in% c("hc", "non_cut")
vst_subset = assay(vst_dds)[, target_samples]
df_anno_subset = as.data.frame(colData(vst_dds)[target_samples, "Subtype", drop=FALSE])


# Distance-based
dist_matrix_subset = as.matrix(dist(t(vst_subset)))

pheatmap(dist_matrix_subset,
         col = colorRampPalette(rev(brewer.pal(9, "Reds")))(255),
         clustering_method = "ward.D2",
         clustering_distance_rows = as.dist(dist_matrix_subset),
         clustering_distance_cols = as.dist(dist_matrix_subset),
         treeheight_col = 0,
         annotation_row = df_anno_subset, annotation_col = df_anno_subset,
         annotation_colors = ann_colors,
         show_colnames = FALSE, show_rownames = FALSE,
         annotation_names_row = FALSE,
         annotation_names_col = FALSE,
         main = "Sample-to-Sample Distance (non cut scl)",
         border_color = NA)

# Correlation-based
cor_matrix_subset = cor(vst_subset, method = "pearson")

pheatmap(cor_matrix_subset,
         breaks = breaks_fixed,
         col = colors_fixed,
         clustering_method = "ward.D2",
         clustering_distance_rows = as.dist(1 - cor_matrix_subset),
         clustering_distance_cols = as.dist(1 - cor_matrix_subset),
         treeheight_col = 0,
         annotation_row = df_anno_subset, annotation_col = df_anno_subset,
         annotation_colors = ann_colors,
         show_colnames = FALSE, show_rownames = FALSE,
         annotation_names_row = FALSE,
         annotation_names_col = FALSE,
         main = "Sample-to-Sample Correlation (non cut scl)",
         border_color = NA)

mean_cor_subset = colMeans(cor_matrix_subset)
sort(mean_cor_subset)[1:10]

## Limited

target_samples = colData(vst_dds)$Subtype %in% c("hc", "limited")
vst_subset = assay(vst_dds)[, target_samples]
df_anno_subset = as.data.frame(colData(vst_dds)[target_samples, "Subtype", drop=FALSE])


# Distance-based
dist_matrix_subset = as.matrix(dist(t(vst_subset)))

pheatmap(dist_matrix_subset,
         col = colorRampPalette(rev(brewer.pal(9, "Reds")))(255),
         clustering_method = "ward.D2",
         clustering_distance_rows = as.dist(dist_matrix_subset),
         clustering_distance_cols = as.dist(dist_matrix_subset),
         treeheight_col = 0,
         annotation_row = df_anno_subset, annotation_col = df_anno_subset,
         annotation_colors = ann_colors,
         show_colnames = FALSE, show_rownames = FALSE,
         annotation_names_row = FALSE,
         annotation_names_col = FALSE,
         main = "Sample-to-Sample Distance (limited scl)",
         border_color = NA)

# Correlation-based
cor_matrix_subset = cor(vst_subset, method = "pearson")

pheatmap(cor_matrix_subset,
         breaks = breaks_fixed,
         col = colors_fixed,
         clustering_method = "ward.D2",
         clustering_distance_rows = as.dist(1 - cor_matrix_subset),
         clustering_distance_cols = as.dist(1 - cor_matrix_subset),
         treeheight_col = 0,
         annotation_row = df_anno_subset, annotation_col = df_anno_subset,
         annotation_colors = ann_colors,
         show_colnames = FALSE, show_rownames = FALSE,
         annotation_names_row = FALSE,
         annotation_names_col = FALSE,
         main = "Sample-to-Sample Correlation (limited scl)",
         border_color = NA)

mean_cor_subset = colMeans(cor_matrix_subset)
sort(mean_cor_subset)[1:10]

## Diffuse

target_samples = colData(vst_dds)$Subtype %in% c("hc", "diffuse")
vst_subset = assay(vst_dds)[, target_samples]
df_anno_subset = as.data.frame(colData(vst_dds)[target_samples, "Subtype", drop=FALSE])


# Distance-based
dist_matrix_subset = as.matrix(dist(t(vst_subset)))

pheatmap(dist_matrix_subset,
         col = colorRampPalette(rev(brewer.pal(9, "Reds")))(255),
         clustering_method = "ward.D2",
         clustering_distance_rows = as.dist(dist_matrix_subset),
         clustering_distance_cols = as.dist(dist_matrix_subset),
         treeheight_col = 0,
         annotation_row = df_anno_subset, annotation_col = df_anno_subset,
         annotation_colors = ann_colors,
         show_colnames = FALSE, show_rownames = FALSE,
         annotation_names_row = FALSE,
         annotation_names_col = FALSE,
         main = "Sample-to-Sample Distance (diffuse scl)",
         border_color = NA)

# Correlation-based
cor_matrix_subset = cor(vst_subset, method = "pearson")

pheatmap(cor_matrix_subset,
         breaks = breaks_fixed,
         col = colors_fixed,
         clustering_method = "ward.D2",
         clustering_distance_rows = as.dist(1 - cor_matrix_subset),
         clustering_distance_cols = as.dist(1 - cor_matrix_subset),
         treeheight_col = 0,
         annotation_row = df_anno_subset, annotation_col = df_anno_subset,
         annotation_colors = ann_colors,
         show_colnames = FALSE, show_rownames = FALSE,
         annotation_names_row = FALSE,
         annotation_names_col = FALSE,
         main = "Sample-to-Sample Correlation (diffuse scl)",
         border_color = NA)

mean_cor_subset = colMeans(cor_matrix_subset)
sort(mean_cor_subset)[1:10]
```

## Final Model Definition / Model-level QC

```{r}
## In this case, we will include one global model and the specific ones

#### Global

# Preliminary model
dds = DESeq(dds)
plotDispEsts(dds, main = "Dispersion Diagnosis - Preliminary")

# Remove outliers
# Aberrant pattern pre and post-analysis
dds = dds[, !dds$Sample %in% c("t19")]

# Final model
dds = DESeq(dds)
plotDispEsts(dds, main = "Dispersion Diagnosis - Final Design")

# Check improvement (now we don't need a blind design)
vst_dds = vst(dds, blind=FALSE)

for (i in c(500,2000,10000)){
  p = plotPCA(vst_dds,
              intgroup = c("Condition", "Subtype"),
              ntop = i,
              pcsToUse = 1:2) +
      geom_text_repel(aes(label = name), size = 3) +
      stat_ellipse(aes(fill = group), geom = "polygon", alpha = 0) +
      scale_color_manual(values = c("skyblue","palegreen3","plum","coral2")) +
      ggtitle(paste0("Top-",i))
  print(p)
}

cor_matrix = cor(assay(vst_dds), method = "pearson")
df_anno = as.data.frame(colData(dds)[, c("Condition", "Subtype")])

pheatmap(cor_matrix,
         breaks = breaks_fixed,
         col = colors_fixed,
         clustering_method = "ward.D2",
         clustering_distance_rows = as.dist(1 - cor_matrix),
         clustering_distance_cols = as.dist(1 - cor_matrix),
         treeheight_col = 0,
         annotation_row = df_anno, annotation_col = df_anno,
         annotation_colors = ann_colors,
         show_colnames = FALSE, show_rownames = FALSE,
         annotation_names_row = FALSE,
         annotation_names_col = FALSE,
         main = "Sample-to-Sample Correlation (VST) - Post-QC",
         border_color = NA)

#### Subtypes

dds_subtypes = dds
design(dds_subtypes) = ~ Subtype
dds_subtypes = DESeq(dds_subtypes)
plotDispEsts(dds_subtypes, main = "Dispersion Diagnosis - Subtypes")
vst_subtypes = vst(dds_subtypes, blind=FALSE)
```

# Differential Expression Analysis (DEA)

## Contrasts and Shrinkage

```{r}
# IMPORTANT: contrast = c("design_column", "target", "reference")

#### Global
res = results(dds, contrast=c("Condition", "scl", "hc"), 
              alpha=0.05)
summary(res)

res_sh = lfcShrink(dds, contrast=c("Condition", "scl", "hc"), 
                   res=res, type="ashr")
summary(res_sh)


#### Subtypes

## Non cut

res_nc = results(dds_subtypes, contrast=c("Subtype", "non_cut", "hc"),
                  alpha=0.05)
summary(res_nc)

res_nc_sh = lfcShrink(dds_subtypes, contrast=c("Subtype", "non_cut", "hc"),
                       res=res_nc, type="ashr")
summary(res_nc_sh)

## Limited

res_lim = results(dds_subtypes, contrast=c("Subtype", "limited", "hc"),
                  alpha=0.05)
summary(res_lim)

res_lim_sh = lfcShrink(dds_subtypes, contrast=c("Subtype", "limited", "hc"),
                       res=res_lim, type="ashr")
summary(res_lim_sh)

## Diffuse

res_dif = results(dds_subtypes, contrast=c("Subtype", "diffuse", "hc"),
                      alpha=0.05)
summary(res_dif)

res_dif_sh = lfcShrink(dds_subtypes, contrast=c("Subtype", "diffuse", "hc"),
                       res=res_dif, type="ashr")
summary(res_dif_sh)
```

## MA plot

```{r}
# Personalized function
deseq_maplot_DEG = function(contrast.res, title_text = "MA-plot: DEGs") {
  
  # Find significant genes without biological threshold
  df = as.data.frame(contrast.res)
  df$sig = ifelse(!is.na(df$padj) & df$padj < 0.05, "Significant", "Not Significant")
  
  # Styling the plot
  ggplot(df, aes(x = baseMean, y = log2FoldChange, color = sig)) +
    geom_point(alpha = 0.5, size = 1.2) +
    scale_x_log10() + 
    scale_color_manual(values = c("Significant" = "coral", 
                                  "Not Significant" = "grey70")) +
    geom_hline(yintercept = 0, color = "black", linetype = "dashed", alpha = 0.5) +
    labs(title = title_text, 
         subtitle = paste("Significant (p-adj <0.05):", sum(df$sig == "Significant")),
         x = "Mean of Normalized Counts", 
         y = "Log2 Fold Change", 
         color = "") +
    theme_minimal() +
    theme(legend.position = "bottom")
}

# Comparing MA plots
deseq_maplot_DEG(res, "MA Plot: scl")
deseq_maplot_DEG(res_sh, "MA Plot: scl (Shrink)")

deseq_maplot_DEG(res_nc,  "MA Plot: non_cut")
deseq_maplot_DEG(res_nc_sh,  "MA Plot: non_cut (Shrink)")

deseq_maplot_DEG(res_lim, "MA Plot: lim")
deseq_maplot_DEG(res_lim_sh, "MA Plot: lim (Shrink)")

deseq_maplot_DEG(res_dif, "MA Plot: dif")
deseq_maplot_DEG(res_dif_sh, "MA Plot: dif (Shrink)")
```

## Volcano plot

```{r}
## FC threshold was changed from 2 to 1.5!!

# Personalized function
deseq_volcano_DEG = function(contrast.res,title_text = "Volcano Plot: DEGs"){
  
  # Find significant and substantial DEGs
  plot_data = as.data.frame(contrast.res)
  
  plot_data$diff_state = "Not significant"
  plot_data$diff_state[plot_data$log2FoldChange > log2(1.5) & 
                         plot_data$padj < 0.05] = "Activated"
  plot_data$diff_state[plot_data$log2FoldChange < -log2(1.5) & 
                         plot_data$padj < 0.05] = "Repressed"

  plot_data$diff_state = factor(plot_data$diff_state, 
                                levels = c("Repressed","Activated","Not significant"))
  
  # Styling the plot
  ggplot(plot_data, aes(x = log2FoldChange, y = -log10(padj), color = diff_state)) +
    geom_point(alpha = 0.5, size = 1.2) +
    scale_color_manual(values = c("Repressed" = "blue", 
                                  "Activated" = "red",
                                  "Not significant" = "grey70")) +
    geom_vline(xintercept = c(-log2(1.5), log2(1.5)), 
               linetype = "dashed", color = "black", alpha = 0.5) +
    geom_hline(yintercept = -log10(0.05), 
               linetype = "dashed", color = "black", alpha = 0.5) +
    labs(title = title_text,
         subtitle = paste("Repressed:", sum(plot_data$diff_state == "Repressed"),
                          "| Activated:", sum(plot_data$diff_state == "Activated")),
         x = "log2 Fold Change",
         y = "-log10 p-adj value",
         color = "") +
    theme_minimal() +
    theme(legend.position = "bottom")
}

# Comparing volcano plots
deseq_volcano_DEG(res, "Volcano Plot: scl")
deseq_volcano_DEG(res_sh, "Volcano Plot: scl (Shrink)")

deseq_volcano_DEG(res_nc,  "Volcano Plot: non_cut")
deseq_volcano_DEG(res_nc_sh,  "Volcano Plot: non_cut (Shrink)")

deseq_volcano_DEG(res_lim, "Volcano Plot: lim")
deseq_volcano_DEG(res_lim_sh, "Volcano Plot: lim (Shrink)")

deseq_volcano_DEG(res_dif, "Volcano Plot: dif")
deseq_volcano_DEG(res_dif_sh, "Volcano Plot: dif (Shrink)")
```

## Top DEGs Heatmap

```{r}
#### Global

# Select top DEGs and check consistency
top_genes_sh = head(order(res_sh$padj), 100)

# Transformed data and centered
mat = assay(vst_dds)[top_genes_sh, ]
mat = mat - rowMeans(mat)

max_val = 1.5 # max(abs(mat))
my_breaks = seq(-max_val, max_val, length.out = 51)

pheatmap(mat, 
         annotation_col = df_anno, annotation_colors = ann_colors,
         annotation_names_col = FALSE,
         show_colnames = F, show_rownames = F,
         cluster_rows = T, clustering_method = "ward.D2",
         main = "Top 100 DEGs: scl",
         breaks = my_breaks,
         color = colorRampPalette(c("blue", "white", "red"))(50))

#### Subtypes

## Non cut

# Select top DEGs and check consistency
top_genes_sh = head(order(res_nc_sh$padj), 100)

# Transformed data and centered
mat = assay(vst_dds)[top_genes_sh, dds$Subtype %in% c("hc", "non_cut")]
mat = mat - rowMeans(mat)

max_val = 1.5 # max(abs(mat))
my_breaks = seq(-max_val, max_val, length.out = 51)

pheatmap(mat, 
         annotation_col = df_anno, annotation_colors = ann_colors,
         annotation_names_col = FALSE,
         show_colnames = T, show_rownames = F,
         cluster_rows = T, clustering_method = "ward.D2",
         main = "Top 100 DEGs: non_cut",
         breaks = my_breaks,
         color = colorRampPalette(c("blue", "white", "red"))(50))

## Limited

# Select top DEGs and check consistency
top_genes_sh = head(order(res_lim_sh$padj), 100)

# Transformed data and centered
mat = assay(vst_dds)[top_genes_sh, dds$Subtype %in% c("hc", "limited")]
mat = mat - rowMeans(mat)

max_val = 1.5 # max(abs(mat))
my_breaks = seq(-max_val, max_val, length.out = 51)

pheatmap(mat, 
         annotation_col = df_anno, annotation_colors = ann_colors,
         annotation_names_col = FALSE,
         show_colnames = T, show_rownames = F,
         cluster_rows = T, clustering_method = "ward.D2",
         main = "Top 100 DEGs: lim",
         breaks = my_breaks,
         color = colorRampPalette(c("blue", "white", "red"))(50))

## Diffuse

# Select top DEGs and check consistency
top_genes_sh = head(order(res_dif_sh$padj), 100)

# Transformed data and centered
mat = assay(vst_dds)[top_genes_sh, dds$Subtype %in% c("hc", "diffuse")]
mat = mat - rowMeans(mat)

max_val = 1.5 # max(abs(mat))
my_breaks = seq(-max_val, max_val, length.out = 51)

pheatmap(mat, 
         annotation_col = df_anno, annotation_colors = ann_colors,
         annotation_names_col = FALSE,
         show_colnames = T, show_rownames = F,
         cluster_rows = T, clustering_method = "ward.D2",
         main = "Top 100 DEGs: dif",
         breaks = my_breaks,
         color = colorRampPalette(c("blue", "white", "red"))(50))
```

# Enrichment Analysis

## Extract DEGs

```{r}
#### Global

# Upregulated DEGs
upreg_deseq = subset(res_sh, padj < 0.05 & log2FoldChange > log2(1.5))
nrow(upreg_deseq)
upreg_genes = rownames(upreg_deseq)

# Downregulated DEGs
downreg_deseq = subset(res_sh, padj < 0.05 & log2FoldChange < -log2(1.5))
nrow(downreg_deseq)
downreg_genes = rownames(downreg_deseq)

#### Subtypes

## Non cut

# Upregulated DEGs
upreg_nc = subset(res_nc_sh, padj < 0.05 & log2FoldChange > log2(1.5))
nrow(upreg_nc)
upreg_genes_nc = rownames(upreg_nc)

# Downregulated DEGs
downreg_nc = subset(res_nc_sh, padj < 0.05 & log2FoldChange < -log2(1.5))
nrow(downreg_nc)
downreg_genes_nc = rownames(downreg_nc)

## Limited

# Upregulated DEGs
upreg_lim = subset(res_lim_sh, padj < 0.05 & log2FoldChange > log2(1.5))
nrow(upreg_lim)
upreg_genes_lim = rownames(upreg_lim)

# Downregulated DEGs
downreg_lim = subset(res_lim_sh, padj < 0.05 & log2FoldChange < -log2(1.5))
nrow(downreg_lim)
downreg_genes_lim = rownames(downreg_lim)

## Diffuse

# Upregulated DEGs
upreg_dif = subset(res_dif_sh, padj < 0.05 & log2FoldChange > log2(1.5))
nrow(upreg_dif)
upreg_genes_dif = rownames(upreg_dif)

# Downregulated DEGs
downreg_dif = subset(res_dif_sh, padj < 0.05 & log2FoldChange < -log2(1.5))
nrow(downreg_dif)
downreg_genes_dif = rownames(downreg_dif)
```

## Gene Ontology Enrichment

```{r}
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)

#### Global

go_upreg  =  enrichGO(gene          = upreg_genes,
                      OrgDb         = org.Hs.eg.db,
                      keyType       = "ENSEMBL",
                      ont           = "BP",
                      pAdjustMethod = "BH",
                      pvalueCutoff  = 0.05,
                      qvalueCutoff  = 0.05)

df_temp = as.data.frame(go_upreg)

barplot(go_upreg,showCategory = 12)
dotplot(go_upreg,showCategory = 12)
treeplot(pairwise_termsim(go_upreg),showCategory = 12)
emapplot(pairwise_termsim(go_upreg),showCategory = 12)
cnetplot(go_upreg,showCategory = 12)

go_downreg  =  enrichGO(gene        = downreg_genes,
                      OrgDb         = org.Hs.eg.db,
                      keyType       = "ENSEMBL",
                      ont           = "BP",
                      pAdjustMethod = "BH",
                      pvalueCutoff  = 0.05,
                      qvalueCutoff  = 0.05)

df_temp = as.data.frame(go_downreg)

barplot(go_downreg,showCategory = 12)
dotplot(go_downreg,showCategory = 12)
#treeplot(pairwise_termsim(go_downreg),showCategory = 12)
emapplot(pairwise_termsim(go_downreg),showCategory = 12)
cnetplot(go_downreg,showCategory = 12)

#### Subtypes

## Non cut

go_upreg  =  enrichGO(gene          = upreg_genes_nc,
                      OrgDb         = org.Hs.eg.db,
                      keyType       = "ENSEMBL",
                      ont           = "BP",
                      pAdjustMethod = "BH",
                      pvalueCutoff  = 0.05,
                      qvalueCutoff  = 0.05)

df_temp = as.data.frame(go_upreg)

barplot(go_upreg,showCategory = 12)
dotplot(go_upreg,showCategory = 12)
treeplot(pairwise_termsim(go_upreg),showCategory = 12)
emapplot(pairwise_termsim(go_upreg),showCategory = 12)
cnetplot(go_upreg,showCategory = 12)

go_downreg  =  enrichGO(gene        = downreg_genes_nc,
                      OrgDb         = org.Hs.eg.db,
                      keyType       = "ENSEMBL",
                      ont           = "BP",
                      pAdjustMethod = "BH",
                      pvalueCutoff  = 0.05,
                      qvalueCutoff  = 0.05)

df_temp = as.data.frame(go_downreg)

barplot(go_downreg,showCategory = 12)
dotplot(go_downreg,showCategory = 12)
treeplot(pairwise_termsim(go_downreg),showCategory = 12)
emapplot(pairwise_termsim(go_downreg),showCategory = 12)
cnetplot(go_downreg,showCategory = 12)

## Limited

go_upreg  =  enrichGO(gene          = upreg_genes_lim,
                      OrgDb         = org.Hs.eg.db,
                      keyType       = "ENSEMBL",
                      ont           = "BP",
                      pAdjustMethod = "BH",
                      pvalueCutoff  = 0.05,
                      qvalueCutoff  = 0.05)

df_temp = as.data.frame(go_upreg)

barplot(go_upreg,showCategory = 12)
dotplot(go_upreg,showCategory = 12)
treeplot(pairwise_termsim(go_upreg),showCategory = 12)
emapplot(pairwise_termsim(go_upreg),showCategory = 12)
cnetplot(go_upreg,showCategory = 12)

# go_downreg  =  enrichGO(gene        = downreg_genes_lim,
#                       OrgDb         = org.Hs.eg.db,
#                       keyType       = "ENSEMBL",
#                       ont           = "BP",
#                       pAdjustMethod = "BH",
#                       pvalueCutoff  = 0.05,
#                       qvalueCutoff  = 0.05)
# 
# df_temp = as.data.frame(go_downreg)
# 
# barplot(go_downreg,showCategory = 12)
# dotplot(go_downreg,showCategory = 12)
# treeplot(pairwise_termsim(go_downreg),showCategory = 12)
# emapplot(pairwise_termsim(go_downreg),showCategory = 12)
# cnetplot(go_downreg,showCategory = 12)

## Diffuse

go_upreg  =  enrichGO(gene          = upreg_genes_dif,
                      OrgDb         = org.Hs.eg.db,
                      keyType       = "ENSEMBL",
                      ont           = "BP",
                      pAdjustMethod = "BH",
                      pvalueCutoff  = 0.05,
                      qvalueCutoff  = 0.05)

df_temp = as.data.frame(go_upreg)

barplot(go_upreg,showCategory = 12)
dotplot(go_upreg,showCategory = 12)
treeplot(pairwise_termsim(go_upreg),showCategory = 12)
emapplot(pairwise_termsim(go_upreg),showCategory = 12)
cnetplot(go_upreg,showCategory = 12)

go_downreg  =  enrichGO(gene        = downreg_genes_dif,
                      OrgDb         = org.Hs.eg.db,
                      keyType       = "ENSEMBL",
                      ont           = "BP",
                      pAdjustMethod = "BH",
                      pvalueCutoff  = 0.05,
                      qvalueCutoff  = 0.05)

df_temp = as.data.frame(go_downreg)

barplot(go_downreg,showCategory = 12)
dotplot(go_downreg,showCategory = 12)
#treeplot(pairwise_termsim(go_downreg),showCategory = 12)
emapplot(pairwise_termsim(go_downreg),showCategory = 12)
cnetplot(go_downreg,showCategory = 12)
```

## KEGG Enrichment

```{r}
#### Global

gene_conv_upreg = bitr(upreg_genes, 
                       fromType = "ENSEMBL", 
                       toType   = "ENTREZID", 
                       OrgDb    = org.Hs.eg.db)

upreg_entrez = gene_conv_upreg$ENTREZID


kegg_upreg = enrichKEGG(gene          = upreg_entrez,
                        organism      = 'hsa',
                        pAdjustMethod = "BH",
                        pvalueCutoff  = 0.05)

df_temp = as.data.frame(kegg_upreg)

barplot(kegg_upreg,showCategory = 12)
dotplot(kegg_upreg,showCategory = 12)
treeplot(pairwise_termsim(kegg_upreg),showCategory = 12)
emapplot(pairwise_termsim(kegg_upreg),showCategory = 12)
cnetplot(kegg_upreg,showCategory = 12)

#### Subtypes

## Non cut

gene_conv_upreg = bitr(upreg_genes_nc, 
                       fromType = "ENSEMBL", 
                       toType   = "ENTREZID", 
                       OrgDb    = org.Hs.eg.db)

upreg_entrez = gene_conv_upreg$ENTREZID


kegg_upreg = enrichKEGG(gene          = upreg_entrez,
                        organism      = 'hsa',
                        pAdjustMethod = "BH",
                        pvalueCutoff  = 0.05)

df_temp = as.data.frame(kegg_upreg)

barplot(kegg_upreg,showCategory = 12)
dotplot(kegg_upreg,showCategory = 12)
treeplot(pairwise_termsim(kegg_upreg),showCategory = 12)
emapplot(pairwise_termsim(kegg_upreg),showCategory = 12)
cnetplot(kegg_upreg,showCategory = 12)

## Limited

gene_conv_upreg = bitr(upreg_genes_lim, 
                       fromType = "ENSEMBL", 
                       toType   = "ENTREZID", 
                       OrgDb    = org.Hs.eg.db)

upreg_entrez = gene_conv_upreg$ENTREZID


kegg_upreg = enrichKEGG(gene          = upreg_entrez,
                        organism      = 'hsa',
                        pAdjustMethod = "BH",
                        pvalueCutoff  = 0.05)

df_temp = as.data.frame(kegg_upreg)

barplot(kegg_upreg,showCategory = 12)
dotplot(kegg_upreg,showCategory = 12)
treeplot(pairwise_termsim(kegg_upreg),showCategory = 12)
emapplot(pairwise_termsim(kegg_upreg),showCategory = 12)
cnetplot(kegg_upreg,showCategory = 12)

## Diffuse

gene_conv_upreg = bitr(upreg_genes_dif, 
                       fromType = "ENSEMBL", 
                       toType   = "ENTREZID", 
                       OrgDb    = org.Hs.eg.db)

upreg_entrez = gene_conv_upreg$ENTREZID


kegg_upreg = enrichKEGG(gene          = upreg_entrez,
                        organism      = 'hsa',
                        pAdjustMethod = "BH",
                        pvalueCutoff  = 0.05)

df_temp = as.data.frame(kegg_upreg)

barplot(kegg_upreg,showCategory = 12)
dotplot(kegg_upreg,showCategory = 12)
#treeplot(pairwise_termsim(kegg_upreg),showCategory = 12)
emapplot(pairwise_termsim(kegg_upreg),showCategory = 12)
cnetplot(kegg_upreg,showCategory = 12)
```

## Pathway Enrichment

```{r,eval=F}
# library(pathview)
# 
# # Make an Ensembl to Entrez conversion
# res_df = as.data.frame(res_pos_sh)
# res_df$ensembl = rownames(res_df)
# 
# ids = bitr(res_df$ensembl, 
#            fromType = "ENSEMBL", 
#            toType   = "ENTREZID", 
#            OrgDb    = org.Hs.eg.db)
# 
# # Pathview needs a named vector (log2FC associated to ENtrezID)
# input_data = merge(res_df, ids, by.x = "ensembl", by.y = "ENSEMBL")
# gene_list = input_data$log2FoldChange
# names(gene_list) = input_data$ENTREZID
# 
# # Pathview example
# pahview(gene.data  = gene_list,
#          pathway.id = "hsa04621", # "NOD-like receptor signaling pathway"
#          species    = "hsa",
#          limit      = list(gene = 1, cpd = 1),
#          low        = "blue",
#          mid        = "gray",
#          high       = "red",)
```

## GeneSet Enrichment

```{r}
#### Global

# Rank genes and omit NAs
ranked_genes = sign(res_sh$log2FoldChange) * -log10(res_sh$pvalue)
names(ranked_genes) = rownames(res_sh)
ranked_genes = sort(na.omit(ranked_genes), decreasing = TRUE)

# GO GSEA
gsea_go = gseGO(geneList     = ranked_genes,
                OrgDb        = org.Hs.eg.db,
                keyType      = "ENSEMBL",
                ont          = "BP",
                minGSSize    = 15,
                maxGSSize    = 500,
                pvalueCutoff = 0.05,
                pAdjustMethod = "BH",
                seed         = 123)

dotplot(gsea_go, showCategory = 12, split = ".sign") + 
  facet_grid(. ~ .sign)

ridgeplot(gsea_go, showCategory = 12)


# KEGG GSEA

# Entrez ID needed
ids_gsea = bitr(names(ranked_genes),
                fromType = "ENSEMBL",
                toType   = "ENTREZID",
                OrgDb    = org.Hs.eg.db)
ranked_entrez = ranked_genes[ids_gsea$ENSEMBL]
names(ranked_entrez) = ids_gsea$ENTREZID
ranked_entrez = ranked_entrez[!duplicated(names(ranked_entrez))]
ranked_entrez = sort(ranked_entrez, decreasing = TRUE)

gsea_kegg = gseKEGG(geneList      = ranked_entrez,
                    organism      = "hsa",
                    minGSSize     = 15,
                    maxGSSize     = 500,
                    pvalueCutoff  = 0.05,
                    pAdjustMethod = "BH",
                    seed          = 123)

dotplot(gsea_kegg, showCategory = 12, split = ".sign") + 
  facet_grid(. ~ .sign)

#### Subtypes

## Non cut

# Rank genes and omit NAs
ranked_genes = sign(res_nc_sh$log2FoldChange) * -log10(res_nc_sh$pvalue)
names(ranked_genes) = rownames(res_nc_sh)
ranked_genes = sort(na.omit(ranked_genes), decreasing = TRUE)

# GO GSEA
gsea_go = gseGO(geneList     = ranked_genes,
                OrgDb        = org.Hs.eg.db,
                keyType      = "ENSEMBL",
                ont          = "BP",
                minGSSize    = 15,
                maxGSSize    = 500,
                pvalueCutoff = 0.05,
                pAdjustMethod = "BH",
                seed         = 123)

dotplot(gsea_go, showCategory = 12, split = ".sign") + 
  facet_grid(. ~ .sign)

ridgeplot(gsea_go, showCategory = 12)


# KEGG GSEA

# Entrez ID needed
ids_gsea = bitr(names(ranked_genes),
                fromType = "ENSEMBL",
                toType   = "ENTREZID",
                OrgDb    = org.Hs.eg.db)
ranked_entrez = ranked_genes[ids_gsea$ENSEMBL]
names(ranked_entrez) = ids_gsea$ENTREZID
ranked_entrez = ranked_entrez[!duplicated(names(ranked_entrez))]
ranked_entrez = sort(ranked_entrez, decreasing = TRUE)

gsea_kegg = gseKEGG(geneList      = ranked_entrez,
                    organism      = "hsa",
                    minGSSize     = 15,
                    maxGSSize     = 500,
                    pvalueCutoff  = 0.05,
                    pAdjustMethod = "BH",
                    seed          = 123)

dotplot(gsea_kegg, showCategory = 12, split = ".sign") + 
  facet_grid(. ~ .sign)

## Limited

# Rank genes and omit NAs
ranked_genes = sign(res_lim_sh$log2FoldChange) * -log10(res_lim_sh$pvalue)
names(ranked_genes) = rownames(res_lim_sh)
ranked_genes = sort(na.omit(ranked_genes), decreasing = TRUE)

# GO GSEA
gsea_go = gseGO(geneList     = ranked_genes,
                OrgDb        = org.Hs.eg.db,
                keyType      = "ENSEMBL",
                ont          = "BP",
                minGSSize    = 15,
                maxGSSize    = 500,
                pvalueCutoff = 0.05,
                pAdjustMethod = "BH",
                seed         = 123)

dotplot(gsea_go, showCategory = 12, split = ".sign") + 
  facet_grid(. ~ .sign)

ridgeplot(gsea_go, showCategory = 12)


# KEGG GSEA

# Entrez ID needed
ids_gsea = bitr(names(ranked_genes),
                fromType = "ENSEMBL",
                toType   = "ENTREZID",
                OrgDb    = org.Hs.eg.db)
ranked_entrez = ranked_genes[ids_gsea$ENSEMBL]
names(ranked_entrez) = ids_gsea$ENTREZID
ranked_entrez = ranked_entrez[!duplicated(names(ranked_entrez))]
ranked_entrez = sort(ranked_entrez, decreasing = TRUE)

gsea_kegg = gseKEGG(geneList      = ranked_entrez,
                    organism      = "hsa",
                    minGSSize     = 15,
                    maxGSSize     = 500,
                    pvalueCutoff  = 0.05,
                    pAdjustMethod = "BH",
                    seed          = 123)

dotplot(gsea_kegg, showCategory = 12, split = ".sign") + 
  facet_grid(. ~ .sign)

## Diffuse

# Rank genes and omit NAs
ranked_genes = sign(res_dif_sh$log2FoldChange) * -log10(res_dif_sh$pvalue)
names(ranked_genes) = rownames(res_dif_sh)
ranked_genes = sort(na.omit(ranked_genes), decreasing = TRUE)

# GO GSEA
gsea_go = gseGO(geneList     = ranked_genes,
                OrgDb        = org.Hs.eg.db,
                keyType      = "ENSEMBL",
                ont          = "BP",
                minGSSize    = 15,
                maxGSSize    = 500,
                pvalueCutoff = 0.05,
                pAdjustMethod = "BH",
                seed         = 123)

dotplot(gsea_go, showCategory = 12, split = ".sign") + 
  facet_grid(. ~ .sign)

ridgeplot(gsea_go, showCategory = 12)


# KEGG GSEA

# Entrez ID needed
ids_gsea = bitr(names(ranked_genes),
                fromType = "ENSEMBL",
                toType   = "ENTREZID",
                OrgDb    = org.Hs.eg.db)
ranked_entrez = ranked_genes[ids_gsea$ENSEMBL]
names(ranked_entrez) = ids_gsea$ENTREZID
ranked_entrez = ranked_entrez[!duplicated(names(ranked_entrez))]
ranked_entrez = sort(ranked_entrez, decreasing = TRUE)

gsea_kegg = gseKEGG(geneList      = ranked_entrez,
                    organism      = "hsa",
                    minGSSize     = 15,
                    maxGSSize     = 500,
                    pvalueCutoff  = 0.05,
                    pAdjustMethod = "BH",
                    seed          = 123)

dotplot(gsea_kegg, showCategory = 12, split = ".sign") + 
  facet_grid(. ~ .sign)
```

## Export information for web tools and integration

```{r}
GSE124073_Utrecht = list(metadata = colData(dds),
                         global = list(dds = dds,
                                       res = as.data.frame(res_sh),
                                       VST = vst_dds),
                         subtypes = list(dds = dds_subtypes,
                                         res_noncut  = as.data.frame(res_nc_sh),
                                         res_limited = as.data.frame(res_lim_sh),
                                         res_diffuse = as.data.frame(res_dif_sh),
                                         VST = vst_subtypes))

saveRDS(object = GSE124073_Utrecht, file = "data/Outputs/GSE124073_Utrecht.rds")
```

------------------------------------------------------------------------

#### Session Info

```{r}
sessionInfo()
```
