---
title: "GSE249641_SLE"
author: "José Manuel García Pérez"
date: "2026-06-20"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

Note: in this study, separate monocyte subtypes are analyzed. They will be preferably treated individually.

# Structures

## Experimental Design / Metadata

```{r}
# Load file
metadata = read.table(file = "data/SLE_GSE249641/GSE249641.txt", 
                      header = T, sep = "\t")

# Available information
colnames(metadata)
rownames(metadata) = metadata$Sample

# Select relevant information
# Just first visits were previously preselected
metadata = metadata[,c("Sample", "Run", "Condition", 
                       "Cell_subtype", "Age", "Gender")]

# Adapt format
metadata$Condition    = factor(metadata$Condition)
metadata$Cell_subtype = factor(metadata$Cell_subtype)
metadata$Gender       = factor(metadata$Gender)

# Info summary
summary(metadata)

#### Subgroups

## Classical
meta_cla = metadata[seq(from=1,to=57,by=3),]
summary(meta_cla)

## Intermediate
meta_int = metadata[seq(from=2,to=57,by=3),]
summary(meta_int)

## Non-classical
meta_non = metadata[seq(from=3,to=57,by=3),]
summary(meta_non)
```

## From transcript to gene level

```{r}
library(rtracklayer)
library(tximport)

  # # Get tx2gene df from the original annotation file used
  # gtf_df = as.data.frame(import("data/Homo_sapiens.GRCh38.115.chr.gtf.gz"))
  # tx2gene = gtf_df[gtf_df$type == "transcript", c("transcript_id", "gene_id")]
  # tx2gene = tx2gene[!duplicated(tx2gene$transcript_id), ]
  # rownames(tx2gene) = NULL
  # saveRDS(tx2gene, "data/tx2gene_Ensembl115.rds")

# Load the tx2gene df
tx2gene = readRDS("data/tx2gene_Ensembl115.rds")
head(tx2gene)

#### Global

# Path to the quantification files
quant_files = file.path("data/SLE_GSE249641/counts/",
                        paste0(metadata$Sample, "_quant.sf"))

# Import and collapse to gene level
txi = tximport(quant_files, type = "salmon",
               tx2gene = tx2gene, ignoreTxVersion = TRUE)

# Check the final number of genes and samples
dim(txi$counts)

#### Subgroups

## Classical

# Path to the quantification files
quant_files = file.path("data/SLE_GSE249641/counts/",
                        paste0(meta_cla$Sample, "_quant.sf"))

# Import and collapse to gene level
txi_cla = tximport(quant_files, type = "salmon",
               tx2gene = tx2gene, ignoreTxVersion = TRUE)

# Check the final number of genes and samples
dim(txi_cla$counts)

## Intermediate

# Path to the quantification files
quant_files = file.path("data/SLE_GSE249641/counts/",
                        paste0(meta_int$Sample, "_quant.sf"))

# Import and collapse to gene level
txi_int = tximport(quant_files, type = "salmon",
               tx2gene = tx2gene, ignoreTxVersion = TRUE)

# Check the final number of genes and samples
dim(txi_int$counts)

## Non-classical

# Path to the quantification files
quant_files = file.path("data/SLE_GSE249641/counts/",
                        paste0(meta_non$Sample, "_quant.sf"))

# Import and collapse to gene level
txi_non = tximport(quant_files, type = "salmon",
               tx2gene = tx2gene, ignoreTxVersion = TRUE)

# Check the final number of genes and samples
dim(txi_non$counts)
```

## The DESeqDataSet

```{r}
library(DESeq2)

#### Global

# Create the first basic DESeqDataSet
dds = DESeqDataSetFromTximport(txi     = txi,
                               colData = metadata,
                               design  = ~ Condition)

# Ensure the reference level
dds$Condition = relevel(x = dds$Condition, ref = "hc")

#### Subgroups

## Classical

# Create the first basic DESeqDataSet
dds_cla = DESeqDataSetFromTximport(txi     = txi_cla,
                               colData = meta_cla,
                               design  = ~ Condition)

# Ensure the reference level
dds_cla$Condition = relevel(x = dds_cla$Condition, ref = "hc")

## Intermediate

# Create the first basic DESeqDataSet
dds_int = DESeqDataSetFromTximport(txi     = txi_int,
                               colData = meta_int,
                               design  = ~ Condition)

# Ensure the reference level
dds_int$Condition = relevel(x = dds_int$Condition, ref = "hc")

## Non-classical

# Create the first basic DESeqDataSet
dds_non = DESeqDataSetFromTximport(txi     = txi_non,
                               colData = meta_non,
                               design  = ~ Condition)

# Ensure the reference level
dds_non$Condition = relevel(x = dds_non$Condition, ref = "hc")
```

# Exploratory Analysis

## Pre-filtering low counts / Gene-level QC

```{r}
# Smallest group?
table(dds$Condition, dds$Cell_subtype)

#### Global

# Keep genes with at least 10 counts for the smallest group size
keep = rowSums(counts(dds) >= 10) >= 7
dds = dds[keep,]

# How many samples are removed?
table(keep)

#### Subgroups

## Classical

# Keep genes with at least 10 counts for the smallest group size
keep = rowSums(counts(dds_cla) >= 10) >= 7
dds_cla = dds_cla[keep,]

# How many samples are removed?
table(keep)

## Intermediate

# Keep genes with at least 10 counts for the smallest group size
keep = rowSums(counts(dds_int) >= 10) >= 7
dds_int = dds_int[keep,]

# How many samples are removed?
table(keep)

## Non-classical

# Keep genes with at least 10 counts for the smallest group size
keep = rowSums(counts(dds_non) >= 10) >= 7
dds_non = dds_non[keep,]

# How many samples are removed?
table(keep)
```

## Sample-level QC

### Data Transformation

```{r}
library("vsn")
library("ggplot2")

#### Global

# "basic" log2(n+1) transformation
ntd_dds = normTransform(dds)

# Variance Stabilizing Transformation (VST)
vst_dds = vst(dds, blind=TRUE)

# Compare effect on the variance
var_raw = meanSdPlot(counts(dds), plot = FALSE)$gg + ggtitle("Raw Counts")
var_raw

var_ntd = meanSdPlot(assay(ntd_dds), plot = FALSE)$gg + ggtitle("log2(n+1)")
var_ntd

var_vst = meanSdPlot(assay(vst_dds), plot = FALSE)$gg + ggtitle("VST")
var_vst

# Final improvement in distribution check
boxplot(counts(dds), outline=T, 
        main ="Raw Data", xaxt="n",
        xlab = paste0("Samples, n=", nrow(metadata)),
        ylab = "Raw Counts",
        col  = ifelse(metadata$Condition == "hc", "skyblue", "salmon"),
        ylim = c(min(counts(dds))-1, max(counts(dds))+3))
legend(x = "topright", legend = c("Control", "Target"), 
       fill = c("skyblue", "salmon"), bty = "n", cex = 0.8)

boxplot(assay(vst_dds), outline=T, 
        main ="VST", xaxt="n",
        xlab = paste0("Samples, n=", nrow(metadata)),
        ylab = "Transformed Counts",
        col  = ifelse(metadata$Condition == "hc", "skyblue", "salmon"),
        ylim = c(min(assay(vst_dds))-1, max(assay(vst_dds))+3))
legend(x = "topright", legend = c("Control", "Target"), 
       fill = c("skyblue", "salmon"), bty = "n", cex = 0.8)

#### Subgroups

## Classical

# "basic" log2(n+1) transformation
ntd_dds_cla = normTransform(dds_cla)

# Variance Stabilizing Transformation (VST)
vst_dds_cla = vst(dds_cla, blind=TRUE)

# Compare effect on the variance
var_raw = meanSdPlot(counts(dds_cla), plot = FALSE)$gg + ggtitle("Raw Counts")
var_raw

var_ntd = meanSdPlot(assay(ntd_dds_cla), plot = FALSE)$gg + ggtitle("log2(n+1)")
var_ntd

var_vst = meanSdPlot(assay(vst_dds_cla), plot = FALSE)$gg + ggtitle("VST")
var_vst

# Final improvement in distribution check
boxplot(counts(dds_cla), outline=T, 
        main ="Raw Data", xaxt="n",
        xlab = paste0("Samples, n=", nrow(meta_cla)),
        ylab = "Raw Counts",
        col  = ifelse(meta_cla$Condition == "hc", "skyblue", "salmon"),
        ylim = c(min(counts(dds_cla))-1, max(counts(dds_cla))+3))
legend(x = "topright", legend = c("Control", "Target"), 
       fill = c("skyblue", "salmon"), bty = "n", cex = 0.8)

boxplot(assay(vst_dds_cla), outline=T, 
        main ="VST", xaxt="n",
        xlab = paste0("Samples, n=", nrow(meta_cla)),
        ylab = "Transformed Counts",
        col  = ifelse(meta_cla$Condition == "hc", "skyblue", "salmon"),
        ylim = c(min(assay(vst_dds_cla))-1, max(assay(vst_dds_cla))+3))
legend(x = "topright", legend = c("Control", "Target"), 
       fill = c("skyblue", "salmon"), bty = "n", cex = 0.8)

## Intermediate

# "basic" log2(n+1) transformation
ntd_dds_int = normTransform(dds_int)

# Variance Stabilizing Transformation (VST)
vst_dds_int = vst(dds_int, blind=TRUE)

# Compare effect on the variance
var_raw = meanSdPlot(counts(dds_int), plot = FALSE)$gg + ggtitle("Raw Counts")
var_raw

var_ntd = meanSdPlot(assay(ntd_dds_int), plot = FALSE)$gg + ggtitle("log2(n+1)")
var_ntd

var_vst = meanSdPlot(assay(vst_dds_int), plot = FALSE)$gg + ggtitle("VST")
var_vst

# Final improvement in distribution check
boxplot(counts(dds_int), outline=T, 
        main ="Raw Data", xaxt="n",
        xlab = paste0("Samples, n=", nrow(meta_int)),
        ylab = "Raw Counts",
        col  = ifelse(meta_int$Condition == "hc", "skyblue", "salmon"),
        ylim = c(min(counts(dds_int))-1, max(counts(dds_int))+3))
legend(x = "topright", legend = c("Control", "Target"), 
       fill = c("skyblue", "salmon"), bty = "n", cex = 0.8)

boxplot(assay(vst_dds_int), outline=T, 
        main ="VST", xaxt="n",
        xlab = paste0("Samples, n=", nrow(meta_int)),
        ylab = "Transformed Counts",
        col  = ifelse(meta_int$Condition == "hc", "skyblue", "salmon"),
        ylim = c(min(assay(vst_dds_int))-1, max(assay(vst_dds_int))+3))
legend(x = "topright", legend = c("Control", "Target"), 
       fill = c("skyblue", "salmon"), bty = "n", cex = 0.8)

## Non-classical

# "basic" log2(n+1) transformation
ntd_dds_non = normTransform(dds_non)

# Variance Stabilizing Transformation (VST)
vst_dds_non = vst(dds_non, blind=TRUE)

# Compare effect on the variance
var_raw = meanSdPlot(counts(dds_non), plot = FALSE)$gg + ggtitle("Raw Counts")
var_raw

var_ntd = meanSdPlot(assay(ntd_dds_non), plot = FALSE)$gg + ggtitle("log2(n+1)")
var_ntd

var_vst = meanSdPlot(assay(vst_dds_non), plot = FALSE)$gg + ggtitle("VST")
var_vst

# Final improvement in distribution check
boxplot(counts(dds_non), outline=T, 
        main ="Raw Data", xaxt="n",
        xlab = paste0("Samples, n=", nrow(meta_non)),
        ylab = "Raw Counts",
        col  = ifelse(meta_non$Condition == "hc", "skyblue", "salmon"),
        ylim = c(min(counts(dds_non))-1, max(counts(dds_non))+3))
legend(x = "topright", legend = c("Control", "Target"), 
       fill = c("skyblue", "salmon"), bty = "n", cex = 0.8)

boxplot(assay(vst_dds_non), outline=T, 
        main ="VST", xaxt="n",
        xlab = paste0("Samples, n=", nrow(meta_non)),
        ylab = "Transformed Counts",
        col  = ifelse(meta_non$Condition == "hc", "skyblue", "salmon"),
        ylim = c(min(assay(vst_dds_non))-1, max(assay(vst_dds_non))+3))
legend(x = "topright", legend = c("Control", "Target"), 
       fill = c("skyblue", "salmon"), bty = "n", cex = 0.8)
```

### Principal Components Analysis (PCA)

```{r}
#### Global

# Condition
plotPCA(vst_dds,
        intgroup = c("Condition"),
        ntop = 500,
        pcsToUse = 1:2) + 
  stat_ellipse(aes(fill = group), geom = "polygon", alpha = 0) +
  scale_color_manual(values = c("skyblue","salmon"))

# Subtype
plotPCA(vst_dds,
        intgroup = c("Cell_subtype"),
        ntop = 500,
        pcsToUse = 1:2) + 
  stat_ellipse(aes(fill = group), geom = "polygon", alpha = 0) +
  scale_color_manual(values = c("skyblue","palegreen3", "plum"))

# Age
plotPCA(vst_dds,
        intgroup = c("Age"),
        ntop = 500,
        pcsToUse = 1:2)

# Gender
plotPCA(vst_dds,
        intgroup = c("Gender"),
        ntop = 500,
        pcsToUse = 1:2) + 
  stat_ellipse(aes(fill = group), geom = "polygon", alpha = 0) +
  scale_color_manual(values = c("skyblue","salmon"))

# Condition / Subtype
plotPCA(vst_dds,
        intgroup = c("Condition","Cell_subtype"),
        ntop = 500,
        pcsToUse = 1:2) + 
  stat_ellipse(aes(fill = group), geom = "polygon", alpha = 0) +
  scale_color_manual(values = c("skyblue","palegreen2","plum",
                                "skyblue4","palegreen4","plum4"))

#### Subgroups

library(ggrepel)

## Classical

for (i in c(500,2000,10000)){
  p = plotPCA(vst_dds_cla,
              intgroup = c("Condition"),
              ntop = i,
              pcsToUse = c(1,2)) + 
      geom_text_repel(aes(label = name), size = 3) +
      stat_ellipse(aes(fill = group), geom = "polygon", alpha = 0) +
      scale_color_manual(values = c("skyblue","salmon")) +
      ggtitle(paste0("Top-",i))
  print(p)
}  

for (i in c(500,2000,10000)){
  p = plotPCA(vst_dds_cla,
              intgroup = c("Condition"),
              ntop = i,
              pcsToUse = c(1,3)) + 
      geom_text_repel(aes(label = name), size = 3) +
      stat_ellipse(aes(fill = group), geom = "polygon", alpha = 0) +
      scale_color_manual(values = c("skyblue","salmon")) +
      ggtitle(paste0("Top-",i))
  print(p)
}

for (i in c(500,2000,10000)){
  p = plotPCA(vst_dds_cla,
              intgroup = c("Condition"),
              ntop = i,
              pcsToUse = c(2,3)) + 
      geom_text_repel(aes(label = name), size = 3) +
      stat_ellipse(aes(fill = group), geom = "polygon", alpha = 0) +
      scale_color_manual(values = c("skyblue","salmon")) +
      ggtitle(paste0("Top-",i))
  print(p)
}

## Intermediate

for (i in c(500,2000,10000)){
  p = plotPCA(vst_dds_int,
              intgroup = c("Condition"),
              ntop = i,
              pcsToUse = c(1,2)) + 
      geom_text_repel(aes(label = name), size = 3) +
      stat_ellipse(aes(fill = group), geom = "polygon", alpha = 0) +
      scale_color_manual(values = c("skyblue","salmon")) +
      ggtitle(paste0("Top-",i))
  print(p)
}  

for (i in c(500,2000,10000)){
  p = plotPCA(vst_dds_int,
              intgroup = c("Condition"),
              ntop = i,
              pcsToUse = c(1,3)) + 
      geom_text_repel(aes(label = name), size = 3) +
      stat_ellipse(aes(fill = group), geom = "polygon", alpha = 0) +
      scale_color_manual(values = c("skyblue","salmon")) +
      ggtitle(paste0("Top-",i))
  print(p)
}

for (i in c(500,2000,10000)){
  p = plotPCA(vst_dds_int,
              intgroup = c("Condition"),
              ntop = i,
              pcsToUse = c(2,3)) + 
      geom_text_repel(aes(label = name), size = 3) +
      stat_ellipse(aes(fill = group), geom = "polygon", alpha = 0) +
      scale_color_manual(values = c("skyblue","salmon")) +
      ggtitle(paste0("Top-",i))
  print(p)
}

## Non-classical

for (i in c(500,2000,10000)){
  p = plotPCA(vst_dds_non,
              intgroup = c("Condition"),
              ntop = i,
              pcsToUse = c(1,2)) + 
      geom_text_repel(aes(label = name), size = 3) +
      stat_ellipse(aes(fill = group), geom = "polygon", alpha = 0) +
      scale_color_manual(values = c("skyblue","salmon")) +
      ggtitle(paste0("Top-",i))
  print(p)
}  

for (i in c(500,2000,10000)){
  p = plotPCA(vst_dds_non,
              intgroup = c("Condition"),
              ntop = i,
              pcsToUse = c(1,3)) + 
      geom_text_repel(aes(label = name), size = 3) +
      stat_ellipse(aes(fill = group), geom = "polygon", alpha = 0) +
      scale_color_manual(values = c("skyblue","salmon")) +
      ggtitle(paste0("Top-",i))
  print(p)
}

for (i in c(500,2000,10000)){
  p = plotPCA(vst_dds_non,
              intgroup = c("Condition"),
              ntop = i,
              pcsToUse = c(2,3)) + 
      geom_text_repel(aes(label = name), size = 3) +
      stat_ellipse(aes(fill = group), geom = "polygon", alpha = 0) +
      scale_color_manual(values = c("skyblue","salmon")) +
      ggtitle(paste0("Top-",i))
  print(p)
}
```

### Sample clustering

```{r}
library(pheatmap)
library(RColorBrewer)

#### Global

# Define groups and colors
df_anno = as.data.frame(metadata[, c("Condition", "Cell_subtype")])

ann_colors = list(
  Condition  = c(hc = "lightblue", 
                 sle = "salmon"),
  Cell_subtype = c(classical = "palegreen3", 
                   intermediate = "burlywood1", 
                   non_classical = "plum")
)

# Distance-based heatmap
dist_matrix = as.matrix(dist(t(assay(vst_dds))))

pheatmap(dist_matrix,
         col = colorRampPalette(rev(brewer.pal(9, "Reds")))(255),
         clustering_method = "ward.D2",
         clustering_distance_rows = as.dist(dist_matrix),
         clustering_distance_cols = as.dist(dist_matrix),
         treeheight_col = 0,
         annotation_row = df_anno, annotation_col = df_anno,
         annotation_colors = ann_colors,
         show_colnames = FALSE, show_rownames = FALSE,
         annotation_names_row = FALSE,
         annotation_names_col = FALSE,
         main = "Sample-to-Sample Distance (VST) - Global",
         border_color = NA)

# Correlation-based heatmap
cor_matrix = cor(assay(vst_dds), method = "pearson")

# Create fixed color fade to compare
n_breaks = 255
breaks_fixed = seq(min(cor_matrix), max(cor_matrix), length.out = n_breaks+1)
colors_fixed = colorRampPalette(brewer.pal(9, "Reds"))(n_breaks)

pheatmap(cor_matrix,
         breaks = breaks_fixed,
         col = colors_fixed,
         clustering_method = "ward.D2",
         clustering_distance_rows = as.dist(1 - cor_matrix),
         clustering_distance_cols = as.dist(1 - cor_matrix),
         treeheight_col = 0,
         annotation_row = df_anno, annotation_col = df_anno,
         annotation_colors = ann_colors,
         show_colnames = FALSE, show_rownames = FALSE,
         annotation_names_row = FALSE,
         annotation_names_col = FALSE,
         main = "Sample-to-Sample Correlation (VST) - Global",
         border_color = NA)

# Find the correlation outlier(s)
mean_cor = colMeans(cor_matrix)
sort(mean_cor)[1:10]

#### Subgroups

## Classical

# Define groups and colors

df_anno = as.data.frame(meta_cla[, c("Condition"),drop=F])

ann_colors = list(
  Condition  = c(hc = "lightblue", 
                 sle = "salmon")
)

# Distance-based heatmap
dist_matrix = as.matrix(dist(t(assay(vst_dds_cla))))

pheatmap(dist_matrix,
         col = colorRampPalette(rev(brewer.pal(9, "Reds")))(255),
         clustering_method = "ward.D2",
         clustering_distance_rows = as.dist(dist_matrix),
         clustering_distance_cols = as.dist(dist_matrix),
         treeheight_col = 0,
         annotation_row = df_anno, annotation_col = df_anno,
         annotation_colors = ann_colors,
         show_colnames = FALSE, show_rownames = FALSE,
         annotation_names_row = FALSE,
         annotation_names_col = FALSE,
         main = "Sample-to-Sample Distance (VST) - Classical",
         border_color = NA)

# Correlation-based heatmap
cor_matrix = cor(assay(vst_dds_cla), method = "pearson")

pheatmap(cor_matrix,
         breaks = breaks_fixed,
         col = colors_fixed,
         clustering_method = "ward.D2",
         clustering_distance_rows = as.dist(1 - cor_matrix),
         clustering_distance_cols = as.dist(1 - cor_matrix),
         treeheight_col = 0,
         annotation_row = df_anno, annotation_col = df_anno,
         annotation_colors = ann_colors,
         show_colnames = T, show_rownames = FALSE,
         annotation_names_row = FALSE,
         annotation_names_col = FALSE,
         main = "Sample-to-Sample Correlation (VST) - Classical",
         border_color = NA)

# Find the correlation outlier(s)
mean_cor = colMeans(cor_matrix)
sort(mean_cor)[1:10]

## Intermediate

# Define groups

df_anno = as.data.frame(meta_int[, c("Condition"),drop=F])

# Distance-based heatmap
dist_matrix = as.matrix(dist(t(assay(vst_dds_int))))

pheatmap(dist_matrix,
         col = colorRampPalette(rev(brewer.pal(9, "Reds")))(255),
         clustering_method = "ward.D2",
         clustering_distance_rows = as.dist(dist_matrix),
         clustering_distance_cols = as.dist(dist_matrix),
         treeheight_col = 0,
         annotation_row = df_anno, annotation_col = df_anno,
         annotation_colors = ann_colors,
         show_colnames = FALSE, show_rownames = FALSE,
         annotation_names_row = FALSE,
         annotation_names_col = FALSE,
         main = "Sample-to-Sample Distance (VST) - Intermediate",
         border_color = NA)

# Correlation-based heatmap
cor_matrix = cor(assay(vst_dds_int), method = "pearson")

pheatmap(cor_matrix,
         breaks = breaks_fixed,
         col = colors_fixed,
         clustering_method = "ward.D2",
         clustering_distance_rows = as.dist(1 - cor_matrix),
         clustering_distance_cols = as.dist(1 - cor_matrix),
         treeheight_col = 0,
         annotation_row = df_anno, annotation_col = df_anno,
         annotation_colors = ann_colors,
         show_colnames = T, show_rownames = FALSE,
         annotation_names_row = FALSE,
         annotation_names_col = FALSE,
         main = "Sample-to-Sample Correlation (VST) - Intermediate",
         border_color = NA)

# Find the correlation outlier(s)
mean_cor = colMeans(cor_matrix)
sort(mean_cor)[1:10]

## Non-classical

# Define groups

df_anno = as.data.frame(meta_non[, c("Condition"),drop=F])

# Distance-based heatmap
dist_matrix = as.matrix(dist(t(assay(vst_dds_non))))

pheatmap(dist_matrix,
         col = colorRampPalette(rev(brewer.pal(9, "Reds")))(255),
         clustering_method = "ward.D2",
         clustering_distance_rows = as.dist(dist_matrix),
         clustering_distance_cols = as.dist(dist_matrix),
         treeheight_col = 0,
         annotation_row = df_anno, annotation_col = df_anno,
         annotation_colors = ann_colors,
         show_colnames = FALSE, show_rownames = FALSE,
         annotation_names_row = FALSE,
         annotation_names_col = FALSE,
         main = "Sample-to-Sample Distance (VST) - Non-classical",
         border_color = NA)

# Correlation-based heatmap
cor_matrix = cor(assay(vst_dds_non), method = "pearson")

pheatmap(cor_matrix,
         breaks = breaks_fixed,
         col = colors_fixed,
         clustering_method = "ward.D2",
         clustering_distance_rows = as.dist(1 - cor_matrix),
         clustering_distance_cols = as.dist(1 - cor_matrix),
         treeheight_col = 0,
         annotation_row = df_anno, annotation_col = df_anno,
         annotation_colors = ann_colors,
         show_colnames = T, show_rownames = FALSE,
         annotation_names_row = FALSE,
         annotation_names_col = FALSE,
         main = "Sample-to-Sample Correlation (VST) - Non-classical",
         border_color = NA)

# Find the correlation outlier(s)
mean_cor = colMeans(cor_matrix)
sort(mean_cor)[1:10]
```

## Final Model Definition / Model-level QC

```{r}
# No relevant modifications needed

# Global model considering subtype
design(dds) = ~ Cell_subtype + Condition
dds = DESeq(dds)
plotDispEsts(dds, main = "Dispersion Diagnosis - Global")
vst_dds = vst(dds, blind=FALSE)

# Individual models
dds_cla = DESeq(dds_cla)
plotDispEsts(dds_cla, main = "Dispersion Diagnosis - Classical")
vst_dds_cla = vst(dds_cla, blind=FALSE)

dds_int = DESeq(dds_int)
plotDispEsts(dds_int, main = "Dispersion Diagnosis - Intermediate")
vst_dds_int = vst(dds_int, blind=FALSE)

dds_non = DESeq(dds_non)
plotDispEsts(dds_non, main = "Dispersion Diagnosis - Non-classical")
vst_dds_non = vst(dds_non, blind=FALSE)
```

# Differential Expression Analysis (DEA)

## Contrasts and Shrinkage

```{r}
# IMPORTANT: contrast = c("design_column", "target", "reference")

# Obtain results for the desired contrasts + Shrink

#### Global

res = results(dds, contrast=c("Condition", "sle", "hc"), 
              alpha=0.05)
summary(res)

res_sh = lfcShrink(dds, contrast=c("Condition", "sle", "hc"), 
                   res=res, type="ashr")
summary(res_sh)

#### Subtypes

## Classical
res_cla = results(dds_cla, contrast=c("Condition", "sle", "hc"), 
                  alpha=0.05)
summary(res_cla)

res_cla_sh = lfcShrink(dds_cla, contrast=c("Condition", "sle", "hc"), 
                       res=res_cla, type="ashr")
summary(res_cla_sh)

## Intermediate
res_int = results(dds_int, contrast=c("Condition", "sle", "hc"), 
                  alpha=0.05)
summary(res_int)

res_int_sh = lfcShrink(dds_int, contrast=c("Condition", "sle", "hc"), 
                       res=res_int, type="ashr")
summary(res_int_sh)

## Non-classical
res_non = results(dds_non, contrast=c("Condition", "sle", "hc"), 
                  alpha=0.05)
summary(res_non)

res_non_sh = lfcShrink(dds_non, contrast=c("Condition", "sle", "hc"), 
                       res=res_non, type="ashr")
summary(res_non_sh)
```

## MA plot

```{r}
# Personalized function
deseq_maplot_DEG = function(contrast.res, title_text = "MA-plot: DEGs") {
  
  # Find significant genes without biological threshold
  df = as.data.frame(contrast.res)
  df$sig = ifelse(!is.na(df$padj) & df$padj < 0.05, "Significant", "Not Significant")
  
  # Styling the plot
  ggplot(df, aes(x = baseMean, y = log2FoldChange, color = sig)) +
    geom_point(alpha = 0.5, size = 1.2) +
    scale_x_log10() + 
    scale_color_manual(values = c("Significant" = "coral", 
                                  "Not Significant" = "grey70")) +
    geom_hline(yintercept = 0, color = "black", linetype = "dashed", alpha = 0.5) +
    labs(title = title_text, 
         subtitle = paste("Significant (p-adj <0.05):", sum(df$sig == "Significant")),
         x = "Mean of Normalized Counts", 
         y = "Log2 Fold Change", 
         color = "") +
    theme_minimal() +
    theme(legend.position = "bottom")
}

# Comparing MA plots
deseq_maplot_DEG(res, "MA Plot: SLE global")
deseq_maplot_DEG(res_sh, "MA Plot: SLE global (Shrink)")

deseq_maplot_DEG(res_cla, "MA Plot: SLE classical")
deseq_maplot_DEG(res_cla_sh, "MA Plot: SLE classical (Shrink)")

deseq_maplot_DEG(res_int, "MA Plot: SLE intermediate")
deseq_maplot_DEG(res_int_sh, "MA Plot: SLE intermediate (Shrink)")

deseq_maplot_DEG(res_non, "MA Plot: SLE non-classical")
deseq_maplot_DEG(res_non_sh, "MA Plot: SLE non-classical (Shrink)")
```

## Volcano plot

```{r}
## FC threshold was changed from 2 to 1.5!!

# Personalized function
deseq_volcano_DEG = function(contrast.res,title_text = "Volcano Plot: DEGs"){
  
  # Find significant and substantial DEGs
  plot_data = as.data.frame(contrast.res)
  
  plot_data$diff_state = "Not significant"
  plot_data$diff_state[plot_data$log2FoldChange > log2(1.5) & 
                         plot_data$padj < 0.05] = "Activated"
  plot_data$diff_state[plot_data$log2FoldChange < -log2(1.5) & 
                         plot_data$padj < 0.05] = "Repressed"

  plot_data$diff_state = factor(plot_data$diff_state, 
                                levels = c("Repressed","Activated","Not significant"))
  
  # Styling the plot
  ggplot(plot_data, aes(x = log2FoldChange, y = -log10(padj), color = diff_state)) +
    geom_point(alpha = 0.5, size = 1.2) +
    scale_color_manual(values = c("Repressed" = "blue", 
                                  "Activated" = "red",
                                  "Not significant" = "grey70")) +
    geom_vline(xintercept = c(-log2(1.5), log2(1.5)), 
               linetype = "dashed", color = "black", alpha = 0.5) +
    geom_hline(yintercept = -log10(0.05), 
               linetype = "dashed", color = "black", alpha = 0.5) +
    labs(title = title_text,
         subtitle = paste("Repressed:", sum(plot_data$diff_state == "Repressed"),
                          "| Activated:", sum(plot_data$diff_state == "Activated")),
         x = "log2 Fold Change",
         y = "-log10 p-adj value",
         color = "") +
    theme_minimal() +
    theme(legend.position = "bottom")
}

# Comparing volcano plots
deseq_volcano_DEG(res, "Volcano Plot: SLE global")
deseq_volcano_DEG(res_sh, "Volcano Plot: SLE global (Shrink)")

deseq_volcano_DEG(res_cla, "Volcano Plot: SLE classical")
deseq_volcano_DEG(res_cla_sh, "Volcano Plot: SLE classical (Shrink)")

deseq_volcano_DEG(res_int, "Volcano Plot: SLE intermediate")
deseq_volcano_DEG(res_int_sh, "Volcano Plot: SLE intermediate (Shrink)")

deseq_volcano_DEG(res_non, "Volcano Plot: SLE non-classical")
deseq_volcano_DEG(res_non_sh, "Volcano Plot: SLE non-classical (Shrink)")
```

## Top DEGs Heatmap

```{r}
#### Global
df_anno = as.data.frame(metadata[, c("Condition"),drop=F])

# Select top DEGs and check consistency
top_genes_sh = head(order(res_sh$padj), 100)

# Transformed data and centered
mat = assay(vst_dds)[top_genes_sh, ]
mat = mat - rowMeans(mat)

max_val = 1.5 # max(abs(mat))
my_breaks = seq(-max_val, max_val, length.out = 51)

pheatmap(mat, 
         annotation_col = df_anno, annotation_colors = ann_colors,
         annotation_names_col = FALSE,
         show_colnames = F, show_rownames = F,
         cluster_rows = T, clustering_method = "ward.D2",
         main = "Top 100 DEGs: global",
         breaks = my_breaks,
         color = colorRampPalette(c("blue", "white", "red"))(50))

#### Subtypes

## Classical
df_anno = as.data.frame(meta_cla[, c("Condition"),drop=F])

# Select top DEGs and check consistency
top_genes_sh = head(order(res_cla_sh$padj), 100)

# Transformed data and centered
mat = assay(vst_dds_cla)[top_genes_sh, ]
mat = mat - rowMeans(mat)

max_val = 1.5 # max(abs(mat))
my_breaks = seq(-max_val, max_val, length.out = 51)

pheatmap(mat, 
         annotation_col = df_anno, annotation_colors = ann_colors,
         annotation_names_col = FALSE,
         show_colnames = F, show_rownames = F,
         cluster_rows = T, clustering_method = "ward.D2",
         main = "Top 100 DEGs: classical",
         breaks = my_breaks,
         color = colorRampPalette(c("blue", "white", "red"))(50))

## Intermediate
df_anno = as.data.frame(meta_int[, c("Condition"),drop=F])

# Select top DEGs and check consistency
top_genes_sh = head(order(res_int_sh$padj), 100)

# Transformed data and centered
mat = assay(vst_dds_int)[top_genes_sh, ]
mat = mat - rowMeans(mat)

max_val = 1.5 # max(abs(mat))
my_breaks = seq(-max_val, max_val, length.out = 51)

pheatmap(mat, 
         annotation_col = df_anno, annotation_colors = ann_colors,
         annotation_names_col = FALSE,
         show_colnames = F, show_rownames = F,
         cluster_rows = T, clustering_method = "ward.D2",
         main = "Top 100 DEGs: intermediate",
         breaks = my_breaks,
         color = colorRampPalette(c("blue", "white", "red"))(50))

## Non-classical
df_anno = as.data.frame(meta_non[, c("Condition"),drop=F])

# Select top DEGs and check consistency
top_genes_sh = head(order(res_non_sh$padj), 100)

# Transformed data and centered
mat = assay(vst_dds_non)[top_genes_sh, ]
mat = mat - rowMeans(mat)

max_val = 1.5 # max(abs(mat))
my_breaks = seq(-max_val, max_val, length.out = 51)

pheatmap(mat, 
         annotation_col = df_anno, annotation_colors = ann_colors,
         annotation_names_col = FALSE,
         show_colnames = F, show_rownames = F,
         cluster_rows = T, clustering_method = "ward.D2",
         main = "Top 100 DEGs: non-classical",
         breaks = my_breaks,
         color = colorRampPalette(c("blue", "white", "red"))(50))


```

# Enrichment Analysis

## Extract DEGs

```{r}
#### Global

# Upregulated DEGs
upreg_global = subset(res_sh, padj < 0.05 & log2FoldChange > log2(1.5))
nrow(upreg_global)
upreg_genes_global = rownames(upreg_global)

# Downregulated DEGs
downreg_global = subset(res_sh, padj < 0.05 & log2FoldChange < -log2(1.5))
nrow(downreg_global)
downreg_genes_global = rownames(downreg_global)

#### Subtypes

## Classical

# Upregulated DEGs
upreg_cla = subset(res_cla_sh, padj < 0.05 & log2FoldChange > log2(1.5))
nrow(upreg_cla)
upreg_genes_cla = rownames(upreg_cla)

# Downregulated DEGs
downreg_cla = subset(res_cla_sh, padj < 0.05 & log2FoldChange < -log2(1.5))
nrow(downreg_cla)
downreg_genes_cla = rownames(downreg_cla)

## Intermediate

# Upregulated DEGs
upreg_int = subset(res_int_sh, padj < 0.05 & log2FoldChange > log2(1.5))
nrow(upreg_int)
upreg_genes_int = rownames(upreg_int)

# Downregulated DEGs
downreg_int = subset(res_int_sh, padj < 0.05 & log2FoldChange < -log2(1.5))
nrow(downreg_int)
downreg_genes_int = rownames(downreg_int)

## Non-classical

# Upregulated DEGs
upreg_non = subset(res_non_sh, padj < 0.05 & log2FoldChange > log2(1.5))
nrow(upreg_non)
upreg_genes_non = rownames(upreg_non)

# Downregulated DEGs
downreg_non = subset(res_non_sh, padj < 0.05 & log2FoldChange < -log2(1.5))
nrow(downreg_non)
downreg_genes_non = rownames(downreg_non)
```

## Gene Ontology Enrichment

```{r}
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)

#### Global

go_upreg  =  enrichGO(gene          = upreg_genes_global,
                      OrgDb         = org.Hs.eg.db,
                      keyType       = "ENSEMBL",
                      ont           = "BP",
                      pAdjustMethod = "BH",
                      pvalueCutoff  = 0.05,
                      qvalueCutoff  = 0.05)

df_temp = as.data.frame(go_upreg)

barplot(go_upreg,showCategory = 12)
dotplot(go_upreg,showCategory = 12)
treeplot(pairwise_termsim(go_upreg),showCategory = 12)
emapplot(pairwise_termsim(go_upreg),showCategory = 12)
cnetplot(go_upreg,showCategory = 12)

go_downreg  =  enrichGO(gene        = downreg_genes_global,
                      OrgDb         = org.Hs.eg.db,
                      keyType       = "ENSEMBL",
                      ont           = "BP",
                      pAdjustMethod = "BH",
                      pvalueCutoff  = 0.05,
                      qvalueCutoff  = 0.05)

df_temp = as.data.frame(go_downreg)

barplot(go_downreg,showCategory = 12)
dotplot(go_downreg,showCategory = 12)
treeplot(pairwise_termsim(go_downreg),showCategory = 12)
emapplot(pairwise_termsim(go_downreg),showCategory = 12)
cnetplot(go_downreg,showCategory = 12)

#### Subtypes

## Classical

go_upreg  =  enrichGO(gene          = upreg_genes_cla,
                      OrgDb         = org.Hs.eg.db,
                      keyType       = "ENSEMBL",
                      ont           = "BP",
                      pAdjustMethod = "BH",
                      pvalueCutoff  = 0.05,
                      qvalueCutoff  = 0.05)

df_temp = as.data.frame(go_upreg)

barplot(go_upreg,showCategory = 12)
dotplot(go_upreg,showCategory = 12)
treeplot(pairwise_termsim(go_upreg),showCategory = 12)
emapplot(pairwise_termsim(go_upreg),showCategory = 12)
cnetplot(go_upreg,showCategory = 12)

go_downreg  =  enrichGO(gene        = downreg_genes_cla,
                      OrgDb         = org.Hs.eg.db,
                      keyType       = "ENSEMBL",
                      ont           = "BP",
                      pAdjustMethod = "BH",
                      pvalueCutoff  = 0.05,
                      qvalueCutoff  = 0.05)

df_temp = as.data.frame(go_downreg)

barplot(go_downreg,showCategory = 12)
dotplot(go_downreg,showCategory = 12)
treeplot(pairwise_termsim(go_downreg),showCategory = 12)
emapplot(pairwise_termsim(go_downreg),showCategory = 12)
cnetplot(go_downreg,showCategory = 12)

## Intermediate

go_upreg  =  enrichGO(gene          = upreg_genes_int,
                      OrgDb         = org.Hs.eg.db,
                      keyType       = "ENSEMBL",
                      ont           = "BP",
                      pAdjustMethod = "BH",
                      pvalueCutoff  = 0.05,
                      qvalueCutoff  = 0.05)

df_temp = as.data.frame(go_upreg)

barplot(go_upreg,showCategory = 12)
dotplot(go_upreg,showCategory = 12)
treeplot(pairwise_termsim(go_upreg),showCategory = 12)
emapplot(pairwise_termsim(go_upreg),showCategory = 12)
cnetplot(go_upreg,showCategory = 12)

go_downreg  =  enrichGO(gene        = downreg_genes_int,
                      OrgDb         = org.Hs.eg.db,
                      keyType       = "ENSEMBL",
                      ont           = "BP",
                      pAdjustMethod = "BH",
                      pvalueCutoff  = 0.05,
                      qvalueCutoff  = 0.05)

df_temp = as.data.frame(go_downreg)

barplot(go_downreg,showCategory = 12)
dotplot(go_downreg,showCategory = 12)
treeplot(pairwise_termsim(go_downreg),showCategory = 12)
emapplot(pairwise_termsim(go_downreg),showCategory = 12)
cnetplot(go_downreg,showCategory = 12)

## Non-classical

go_upreg  =  enrichGO(gene          = upreg_genes_non,
                      OrgDb         = org.Hs.eg.db,
                      keyType       = "ENSEMBL",
                      ont           = "BP",
                      pAdjustMethod = "BH",
                      pvalueCutoff  = 0.05,
                      qvalueCutoff  = 0.05)

df_temp = as.data.frame(go_upreg)

barplot(go_upreg,showCategory = 12)
dotplot(go_upreg,showCategory = 12)
treeplot(pairwise_termsim(go_upreg),showCategory = 12)
emapplot(pairwise_termsim(go_upreg),showCategory = 12)
cnetplot(go_upreg,showCategory = 12)

go_downreg  =  enrichGO(gene        = downreg_genes_non,
                      OrgDb         = org.Hs.eg.db,
                      keyType       = "ENSEMBL",
                      ont           = "BP",
                      pAdjustMethod = "BH",
                      pvalueCutoff  = 0.05,
                      qvalueCutoff  = 0.05)

df_temp = as.data.frame(go_downreg)

barplot(go_downreg,showCategory = 12)
dotplot(go_downreg,showCategory = 12)
treeplot(pairwise_termsim(go_downreg),showCategory = 12)
emapplot(pairwise_termsim(go_downreg),showCategory = 12)
cnetplot(go_downreg,showCategory = 12)
```

## KEGG Enrichment

```{r}
#### Global

gene_conv_upreg = bitr(upreg_genes_global, 
                       fromType = "ENSEMBL", 
                       toType   = "ENTREZID", 
                       OrgDb    = org.Hs.eg.db)

upreg_entrez = gene_conv_upreg$ENTREZID


kegg_upreg = enrichKEGG(gene          = upreg_entrez,
                        organism      = 'hsa',
                        pAdjustMethod = "BH",
                        pvalueCutoff  = 0.05)

df_temp = as.data.frame(kegg_upreg)

barplot(kegg_upreg,showCategory = 12)
dotplot(kegg_upreg,showCategory = 12)
treeplot(pairwise_termsim(kegg_upreg),showCategory = 12)
emapplot(pairwise_termsim(kegg_upreg),showCategory = 12)
cnetplot(kegg_upreg,showCategory = 12)

#### Subtypes

## Classical

gene_conv_upreg = bitr(upreg_genes_cla, 
                       fromType = "ENSEMBL", 
                       toType   = "ENTREZID", 
                       OrgDb    = org.Hs.eg.db)

upreg_entrez = gene_conv_upreg$ENTREZID


kegg_upreg = enrichKEGG(gene          = upreg_entrez,
                        organism      = 'hsa',
                        pAdjustMethod = "BH",
                        pvalueCutoff  = 0.05)

df_temp = as.data.frame(kegg_upreg)

barplot(kegg_upreg,showCategory = 12)
dotplot(kegg_upreg,showCategory = 12)
treeplot(pairwise_termsim(kegg_upreg),showCategory = 12)
emapplot(pairwise_termsim(kegg_upreg),showCategory = 12)
cnetplot(kegg_upreg,showCategory = 12)

## Intermediate

gene_conv_upreg = bitr(upreg_genes_int, 
                       fromType = "ENSEMBL", 
                       toType   = "ENTREZID", 
                       OrgDb    = org.Hs.eg.db)

upreg_entrez = gene_conv_upreg$ENTREZID


kegg_upreg = enrichKEGG(gene          = upreg_entrez,
                        organism      = 'hsa',
                        pAdjustMethod = "BH",
                        pvalueCutoff  = 0.05)

df_temp = as.data.frame(kegg_upreg)

barplot(kegg_upreg,showCategory = 12)
dotplot(kegg_upreg,showCategory = 12)
treeplot(pairwise_termsim(kegg_upreg),showCategory = 12)
emapplot(pairwise_termsim(kegg_upreg),showCategory = 12)
cnetplot(kegg_upreg,showCategory = 12)

## Non-classical

gene_conv_upreg = bitr(upreg_genes_non, 
                       fromType = "ENSEMBL", 
                       toType   = "ENTREZID", 
                       OrgDb    = org.Hs.eg.db)

upreg_entrez = gene_conv_upreg$ENTREZID


kegg_upreg = enrichKEGG(gene          = upreg_entrez,
                        organism      = 'hsa',
                        pAdjustMethod = "BH",
                        pvalueCutoff  = 0.05)

df_temp = as.data.frame(kegg_upreg)

barplot(kegg_upreg,showCategory = 12)
dotplot(kegg_upreg,showCategory = 12)
treeplot(pairwise_termsim(kegg_upreg),showCategory = 12)
emapplot(pairwise_termsim(kegg_upreg),showCategory = 12)
cnetplot(kegg_upreg,showCategory = 12)
```

## Pathway Enrichment

```{r, eval=F}
###
```

## GeneSet Enrichment

```{r}
#### Global

# Rank genes and omit NAs
ranked_genes = sign(res_sh$log2FoldChange) * -log10(res_sh$pvalue)
names(ranked_genes) = rownames(res_sh)
ranked_genes = sort(na.omit(ranked_genes), decreasing = TRUE)

# GO GSEA
gsea_go = gseGO(geneList     = ranked_genes,
                OrgDb        = org.Hs.eg.db,
                keyType      = "ENSEMBL",
                ont          = "BP",
                minGSSize    = 15,
                maxGSSize    = 500,
                pvalueCutoff = 0.05,
                pAdjustMethod = "BH",
                seed         = 123)

dotplot(gsea_go, showCategory = 12, split = ".sign") + 
  facet_grid(. ~ .sign)

ridgeplot(gsea_go, showCategory = 12)


# KEGG GSEA

# Entrez ID needed
ids_gsea = bitr(names(ranked_genes),
                fromType = "ENSEMBL",
                toType   = "ENTREZID",
                OrgDb    = org.Hs.eg.db)
ranked_entrez = ranked_genes[ids_gsea$ENSEMBL]
names(ranked_entrez) = ids_gsea$ENTREZID
ranked_entrez = ranked_entrez[!duplicated(names(ranked_entrez))]
ranked_entrez = sort(ranked_entrez, decreasing = TRUE)

gsea_kegg = gseKEGG(geneList      = ranked_entrez,
                    organism      = "hsa",
                    minGSSize     = 15,
                    maxGSSize     = 500,
                    pvalueCutoff  = 0.05,
                    pAdjustMethod = "BH",
                    seed          = 123)

dotplot(gsea_kegg, showCategory = 12, split = ".sign") + 
  facet_grid(. ~ .sign)

#### Subtypes

## Classical

# Rank genes and omit NAs
ranked_genes = sign(res_cla_sh$log2FoldChange) * -log10(res_cla_sh$pvalue)
names(ranked_genes) = rownames(res_cla_sh)
ranked_genes = sort(na.omit(ranked_genes), decreasing = TRUE)

# GO GSEA
gsea_go = gseGO(geneList     = ranked_genes,
                OrgDb        = org.Hs.eg.db,
                keyType      = "ENSEMBL",
                ont          = "BP",
                minGSSize    = 15,
                maxGSSize    = 500,
                pvalueCutoff = 0.05,
                pAdjustMethod = "BH",
                seed         = 123)

dotplot(gsea_go, showCategory = 12, split = ".sign") + 
  facet_grid(. ~ .sign)

ridgeplot(gsea_go, showCategory = 12)


# KEGG GSEA

# Entrez ID needed
ids_gsea = bitr(names(ranked_genes),
                fromType = "ENSEMBL",
                toType   = "ENTREZID",
                OrgDb    = org.Hs.eg.db)
ranked_entrez = ranked_genes[ids_gsea$ENSEMBL]
names(ranked_entrez) = ids_gsea$ENTREZID
ranked_entrez = ranked_entrez[!duplicated(names(ranked_entrez))]
ranked_entrez = sort(ranked_entrez, decreasing = TRUE)

gsea_kegg = gseKEGG(geneList      = ranked_entrez,
                    organism      = "hsa",
                    minGSSize     = 15,
                    maxGSSize     = 500,
                    pvalueCutoff  = 0.05,
                    pAdjustMethod = "BH",
                    seed          = 123)

dotplot(gsea_kegg, showCategory = 12, split = ".sign") + 
  facet_grid(. ~ .sign)

## Intermediate

# Rank genes and omit NAs
ranked_genes = sign(res_int_sh$log2FoldChange) * -log10(res_int_sh$pvalue)
names(ranked_genes) = rownames(res_int_sh)
ranked_genes = sort(na.omit(ranked_genes), decreasing = TRUE)

# GO GSEA
gsea_go = gseGO(geneList     = ranked_genes,
                OrgDb        = org.Hs.eg.db,
                keyType      = "ENSEMBL",
                ont          = "BP",
                minGSSize    = 15,
                maxGSSize    = 500,
                pvalueCutoff = 0.05,
                pAdjustMethod = "BH",
                seed         = 123)

dotplot(gsea_go, showCategory = 12, split = ".sign") + 
  facet_grid(. ~ .sign)

ridgeplot(gsea_go, showCategory = 12)


# KEGG GSEA

# Entrez ID needed
ids_gsea = bitr(names(ranked_genes),
                fromType = "ENSEMBL",
                toType   = "ENTREZID",
                OrgDb    = org.Hs.eg.db)
ranked_entrez = ranked_genes[ids_gsea$ENSEMBL]
names(ranked_entrez) = ids_gsea$ENTREZID
ranked_entrez = ranked_entrez[!duplicated(names(ranked_entrez))]
ranked_entrez = sort(ranked_entrez, decreasing = TRUE)

gsea_kegg = gseKEGG(geneList      = ranked_entrez,
                    organism      = "hsa",
                    minGSSize     = 15,
                    maxGSSize     = 500,
                    pvalueCutoff  = 0.05,
                    pAdjustMethod = "BH",
                    seed          = 123)

dotplot(gsea_kegg, showCategory = 12, split = ".sign") + 
  facet_grid(. ~ .sign)

## Non-classical

# Rank genes and omit NAs
ranked_genes = sign(res_non_sh$log2FoldChange) * -log10(res_non_sh$pvalue)
names(ranked_genes) = rownames(res_non_sh)
ranked_genes = sort(na.omit(ranked_genes), decreasing = TRUE)

# GO GSEA
gsea_go = gseGO(geneList     = ranked_genes,
                OrgDb        = org.Hs.eg.db,
                keyType      = "ENSEMBL",
                ont          = "BP",
                minGSSize    = 15,
                maxGSSize    = 500,
                pvalueCutoff = 0.05,
                pAdjustMethod = "BH",
                seed         = 123)

dotplot(gsea_go, showCategory = 12, split = ".sign") + 
  facet_grid(. ~ .sign)

ridgeplot(gsea_go, showCategory = 12)


# KEGG GSEA

# Entrez ID needed
ids_gsea = bitr(names(ranked_genes),
                fromType = "ENSEMBL",
                toType   = "ENTREZID",
                OrgDb    = org.Hs.eg.db)
ranked_entrez = ranked_genes[ids_gsea$ENSEMBL]
names(ranked_entrez) = ids_gsea$ENTREZID
ranked_entrez = ranked_entrez[!duplicated(names(ranked_entrez))]
ranked_entrez = sort(ranked_entrez, decreasing = TRUE)

gsea_kegg = gseKEGG(geneList      = ranked_entrez,
                    organism      = "hsa",
                    minGSSize     = 15,
                    maxGSSize     = 500,
                    pvalueCutoff  = 0.05,
                    pAdjustMethod = "BH",
                    seed          = 123)

dotplot(gsea_kegg, showCategory = 12, split = ".sign") + 
  facet_grid(. ~ .sign)
```

## Export information for web tools and integration

```{r}
GSE249641 = list(global = list(metadata = colData(dds),
                               dds = dds,
                               res = as.data.frame(res_sh),
                               VST = vst_dds),
                 classical = list(metadata = meta_cla,
                                  dds = dds_cla,
                                  res = as.data.frame(res_cla_sh),
                                  VST = vst_dds_cla),
                 intermediate = list(metadata = meta_int,
                                     dds = dds_int,
                                     res = as.data.frame(res_int_sh),
                                     VST = vst_dds_int),
                 non_classical = list(metadata = meta_non,
                                      dds = dds_non,
                                      res = as.data.frame(res_non_sh),
                                      VST = vst_dds_non))


saveRDS(object = GSE249641, file = "data/Outputs/GSE249641.rds")
```

------------------------------------------------------------------------

#### Session Info

```{r}
sessionInfo()
```
