---
title: "Integrative Analysis"
author: "José Manuel García Pérez"
date: "2026-06-20"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

# Introduction

In the previous notebooks, individual RNA-seq datasets spanning several autoimmune diseases were pre-processed, quantified and analyzed under a unified pipeline. The primary goal of this integrative notebook is to identify an Autoimmunity Core Signature, a set of genes consistently dysregulated across these distinct systemic conditions.

To achieve this, a two-tier approach is implemented:

1.  **Descriptive Overlap Analysis.** A qualitative approach checking the strict intersection of Differentially Expressed Genes (DEGs) across conditions to visualize shared features.

2.  **Statistical Meta-Analysis.** A quantitative integration using Random Effects Models to combine fold changes and p-values, uncovering robust signals that might be missed by simple overlap thresholds.

### Load Results

From each individual analysis, the relevant metadata, DESeqDataSets, VST data and final results were saved using a unified format.

```{r}
# Systemic Lupus Erythematosus
SLE_GSE149050 = readRDS("data/Outputs/GSE149050.rds")
SLE_GSE249641 = readRDS("data/Outputs/GSE249641.rds")

# Systemic Sclerosis / Scleroderma
Scl_GSE303431 = readRDS("data/Outputs/GSE303431.rds")
Scl_GSE124073 = list(Milan   = readRDS("data/Outputs/GSE124073_Milan.rds"),
                     Utrecht = readRDS("data/Outputs/GSE124073_Utrecht.rds"))

# Sjögren's Syndrome & Non-Sjögren's Sicca
SS_GSE173670 = readRDS("data/Outputs/GSE173670.rds")

# Rheumatoid Arthritis
RA_GSE294225 = readRDS("data/Outputs/GSE294225.rds")

# Juvenile Dermatomyositis
JD_GSE221091 = readRDS("data/Outputs/GSE221091.rds")
```

# Descriptive Analysis: DEG Overlap

### Extract all significant and sustantial DEGs

Before performing complex statistical integrations, it is crucial to visually explore the results, looking for shared DEGs over different conditions. We maintain the same thresholds used: adjusted p-value \<0.05 and fold-change \>1.5x. Crucially, we must split the extraction into upregulated and downregulated genes to avoid false shared features.

```{r}
# Extract DEGs function
extract_DEGs = function(res_df, direction, padj_th = 0.05, lfc_th = log2(1.5)) {
  
  # Remove NAs
  res_clean = res_df[!is.na(res_df$padj), ]
  
  # Directional filtering
  if (direction == "up") {
    genes = rownames(res_clean[res_clean$padj < padj_th & 
                               res_clean$log2FoldChange > lfc_th, ])
  } else if (direction == "down") {
    genes = rownames(res_clean[res_clean$padj < padj_th & 
                               res_clean$log2FoldChange < -lfc_th, ])
  }
  return(genes)
}

## Upregulated
upreg_list = list(
  
  # Systemic Lupus Erythematosus
  SLE_IFNpos = extract_DEGs(SLE_GSE149050$res_SLE_IFNpos, "up"),
  SLE_IFNneg = extract_DEGs(SLE_GSE149050$res_SLE_IFNneg, "up"),
  
  SLE = extract_DEGs(SLE_GSE249641$global$res, "up"),
  SLE_Classical = extract_DEGs(SLE_GSE249641$classical$res, "up"),
  SLE_Intermediate = extract_DEGs(SLE_GSE249641$intermediate$res, "up"),
  SLE_NonClassical = extract_DEGs(SLE_GSE249641$non_classical$res, "up"),
  
  # Systemic Sclerosis / Scleroderma
  Scl = extract_DEGs(Scl_GSE303431$global$res, "up"),
  Scl_Limited = extract_DEGs(Scl_GSE303431$subtypes$res_limited, "up"),
  Scl_Diffuse = extract_DEGs(Scl_GSE303431$subtypes$res_diffuse, "up"),
  
  NFScl = extract_DEGs(Scl_GSE124073$Milan$global$res, "up"),
  NFScl_Early = extract_DEGs(Scl_GSE124073$Milan$subtypes$res_early, "up"),
  NFScl_NonCutaneous = extract_DEGs(Scl_GSE124073$Milan$subtypes$res_noncut, "up"),
  
  DefScl = extract_DEGs(Scl_GSE124073$Utrecht$global$res, "up"),
  DefScl_NonCutaneous = extract_DEGs(Scl_GSE124073$Utrecht$subtypes$res_noncut, "up"),
  DefScl_Limited = extract_DEGs(Scl_GSE124073$Utrecht$subtypes$res_limited, "up"),
  DefScl_Diffuse = extract_DEGs(Scl_GSE124073$Utrecht$subtypes$res_diffuse, "up"),
  
  # Sjögren's
  Sjogren = extract_DEGs(SS_GSE173670$res_pSS, "up"),
  
  # Non-Sjögren's Sicca
  SiccaNS = extract_DEGs(SS_GSE173670$res_nS, "up"),
  
  # Rheumatoid Arthritis
  RheumArthritis = extract_DEGs(RA_GSE294225$res, "up"),
  
  # Juvenile Dermatomyositis
  JuvDermatomyositis = extract_DEGs(JD_GSE221091$res, "up")
)

## Downregulated
downreg_list = list(
  
  # Systemic Lupus Erythematosus
  SLE_IFNpos = extract_DEGs(SLE_GSE149050$res_SLE_IFNpos, "down"),
  SLE_IFNneg = extract_DEGs(SLE_GSE149050$res_SLE_IFNneg, "down"),
  
  SLE = extract_DEGs(SLE_GSE249641$global$res, "down"),
  SLE_Classical = extract_DEGs(SLE_GSE249641$classical$res, "down"),
  SLE_Intermediate = extract_DEGs(SLE_GSE249641$intermediate$res, "down"),
  SLE_NonClassical = extract_DEGs(SLE_GSE249641$non_classical$res, "down"),
  
  # Systemic Sclerosis / Scleroderma
  Scl = extract_DEGs(Scl_GSE303431$global$res, "down"),
  Scl_Limited = extract_DEGs(Scl_GSE303431$subtypes$res_limited, "down"),
  Scl_Diffuse = extract_DEGs(Scl_GSE303431$subtypes$res_diffuse, "down"),
  
  NFScl = extract_DEGs(Scl_GSE124073$Milan$global$res, "down"),
  NFScl_Early = extract_DEGs(Scl_GSE124073$Milan$subtypes$res_early, "down"),
  NFScl_NonCutaneous = extract_DEGs(Scl_GSE124073$Milan$subtypes$res_noncut, "down"),
  
  DefScl = extract_DEGs(Scl_GSE124073$Utrecht$global$res, "down"),
  DefScl_NonCutaneous = extract_DEGs(Scl_GSE124073$Utrecht$subtypes$res_noncut, "down"),
  DefScl_Limited = extract_DEGs(Scl_GSE124073$Utrecht$subtypes$res_limited, "down"),
  DefScl_Diffuse = extract_DEGs(Scl_GSE124073$Utrecht$subtypes$res_diffuse, "down"),
  
  # Sjögren's
  Sjogren = extract_DEGs(SS_GSE173670$res_pSS, "down"),
  
  # Non-Sjögren's Sicca
  SiccaNS = extract_DEGs(SS_GSE173670$res_nS, "down"),
  
  # Rheumatoid Arthritis
  RheumArthritis = extract_DEGs(RA_GSE294225$res, "down"),
  
  # Juvenile Dermatomyositis
  JuvDermatomyositis = extract_DEGs(JD_GSE221091$res, "down")
)
```

### Intra-disease Overlap

Previous to looking across different diseases, we must understand the internal structure of each one when different cohorts exist (SLE and sclerosis). This allows us to understand global vs subtype-specific analysis differences (which subtype dominates or does not participate, how similar they are, how the global design increases noise or is actually representative...) or inter-study behavior (large or unexpected lack of overlap, inconsistent amounts of DEGs...).

```{r}
library(ggVennDiagram)
library(ggplot2)

##### UPREG DEGs

# Analyze differences in the amount of DEGs
lapply(upreg_list, length)

### Systemic Lupus Erythematosus

ggVennDiagram(list(SLE_IFNneg = upreg_list$SLE_IFNneg, 
                   SLE_IFNpos = upreg_list$SLE_IFNpos
)) # Lack of IFNneg DEGs

ggVennDiagram(list(Global = upreg_list$SLE,
                   Classical = upreg_list$SLE_Classical,
                   Intermediate = upreg_list$SLE_Intermediate,
                   NonClassical = upreg_list$SLE_NonClassical
)) # 382 noisy DEGs but big overlap

ggVennDiagram(list(SLE_IFN = unique(c(upreg_list$SLE_IFNneg, 
                            upreg_list$SLE_IFNpos)),
                   SLE_subtypes = unique(c(upreg_list$SLE_Classical,
                            upreg_list$SLE_Intermediate,
                            upreg_list$SLE_NonClassical))
)) # Not as much overlap as expected

### Systemic Sclerosis / Scleroderma

ggVennDiagram(list(Global = upreg_list$Scl,
                   Limited = upreg_list$Scl_Limited,
                   Diffuse = upreg_list$Scl_Diffuse
)) # Extreme lack of DEGs

ggVennDiagram(list(Global = upreg_list$NFScl,
                   Early = upreg_list$NFScl_Early,
                   NonCut = upreg_list$NFScl_NonCutaneous
)) # Non Cutaneous disease explains the DEGs

ggVennDiagram(list(Global = upreg_list$DefScl,
                   NonCut = upreg_list$DefScl_NonCutaneous,
                   Limited = upreg_list$DefScl_Limited,
                   Diffuse = upreg_list$DefScl_Diffuse
)) # The subtypes explain the DEGs without clear overlap

ggVennDiagram(list(Scl = unique(c(upreg_list$Scl,
                            upreg_list$Scl_Limited,
                            upreg_list$Scl_Diffuse)),
                   NFScl = unique(c(upreg_list$NFScl,
                            upreg_list$NFScl_Early,
                            upreg_list$NFScl_NonCutaneous)),
                   DefScl = unique(c(upreg_list$DefScl,
                            upreg_list$DefScl_NonCutaneous,
                            upreg_list$DefScl_Limited,
                            upreg_list$DefScl_Diffuse))
)) # Medium but interesting overlap according to biology



##### DOWNREG DEGs

# Analyze differences in the amount of DEGs
lapply(downreg_list, length)

### Systemic Lupus Erythematosus

ggVennDiagram(list(SLE_IFNneg = downreg_list$SLE_IFNneg, 
                   SLE_IFNpos = downreg_list$SLE_IFNpos
)) # Lack of IFNneg DEGs

ggVennDiagram(list(Global = downreg_list$SLE,
                   Classical = downreg_list$SLE_Classical,
                   Intermediate = downreg_list$SLE_Intermediate,
                   NonClassical = downreg_list$SLE_NonClassical
)) # Similar noisy genes and overlap

ggVennDiagram(list(SLE_IFN = unique(c(downreg_list$SLE_IFNneg, 
                            downreg_list$SLE_IFNpos)),
                   SLE_subtypes = unique(c(downreg_list$SLE_Classical,
                            downreg_list$SLE_Intermediate,
                            downreg_list$SLE_NonClassical))
)) # Not as much overlap as expected

### Systemic Sclerosis / Scleroderma

ggVennDiagram(list(Global = downreg_list$Scl,
                   Limited = downreg_list$Scl_Limited,
                   Diffuse = downreg_list$Scl_Diffuse
)) # Extreme lack of DEGs

ggVennDiagram(list(Global = downreg_list$NFScl,
                   Early = downreg_list$NFScl_Early,
                   NonCut = downreg_list$NFScl_NonCutaneous
)) # Extreme lack of DEGs

ggVennDiagram(list(Global = downreg_list$DefScl,
                   NonCut = downreg_list$DefScl_NonCutaneous,
                   Limited = downreg_list$DefScl_Limited,
                   Diffuse = downreg_list$DefScl_Diffuse
)) # NonCut has more DEGs but they are not captured by global design

ggVennDiagram(list(Scl = unique(c(downreg_list$Scl,
                            downreg_list$Scl_Limited,
                            downreg_list$Scl_Diffuse)),
                   NFScl = unique(c(downreg_list$NFScl,
                            downreg_list$NFScl_Early,
                            downreg_list$NFScl_NonCutaneous)),
                   DefScl = unique(c(downreg_list$DefScl,
                            downreg_list$DefScl_NonCutaneous,
                            downreg_list$DefScl_Limited,
                            downreg_list$DefScl_Diffuse))
)) # No overlap
```

### Inter-disease Overlap

Now, we reach the core question: What is shared across all 5 autoimmune diseases? We need to consolidate our lists first, creating a merged list so that each disease is represented by a single, comprehensive list of DEGs. For sclerosis, the unique union of the disease subtypes is considered. However, for the monocyte subtypes in SLE, only the intersection is included (not the global model as too much noise is added), as it is the most similar approach to the rest of the studies where this level of resolution was not achieved.

While we can draw Venn diagrams for 5 sets, it becomes geometrically unreadable. UpSet plots use a matrix layout where:

-   The dots and lines at the bottom show which specific sets are intersecting.
-   The vertical bars show exactly how many genes are in that specific intersection.
-   The horizontal bars show the total size of each disease's DEG list.

```{r}
library(UpSetR)

### Merged lists
merged_upreg_list = list(SLE = 
                         unique(c(upreg_list$SLE_IFNneg, 
                                  upreg_list$SLE_IFNpos,
                                  Reduce(intersect, list(
                                         upreg_list$SLE_Classical,
                                         upreg_list$SLE_Intermediate,
                                         upreg_list$SLE_NonClassical)))),
                         Sclerosis =
                         unique(c(upreg_list$Scl,
                                  upreg_list$Scl_Limited,
                                  upreg_list$Scl_Diffuse,
                                  upreg_list$NFScl,
                                  upreg_list$NFScl_Early,
                                  upreg_list$NFScl_NonCutaneous,
                                  upreg_list$DefScl,
                                  upreg_list$DefScl_NonCutaneous,
                                  upreg_list$DefScl_Limited,
                                  upreg_list$DefScl_Diffuse)),
                         Sjogren = 
                         upreg_list$Sjogren,
                         Arthritis = 
                         upreg_list$RheumArthritis,
                         Dermatomyositis = 
                         upreg_list$JuvDermatomyositis
)

merged_downreg_list = list(SLE = 
                         unique(c(downreg_list$SLE_IFNneg, 
                                  downreg_list$SLE_IFNpos,
                                  Reduce(intersect, list(
                                         downreg_list$SLE_Classical,
                                         downreg_list$SLE_Intermediate,
                                         downreg_list$SLE_NonClassical)))),
                         Sclerosis =
                         unique(c(downreg_list$Scl,
                                  downreg_list$Scl_Limited,
                                  downreg_list$Scl_Diffuse,
                                  downreg_list$NFScl,
                                  downreg_list$NFScl_Early,
                                  downreg_list$NFScl_NonCutaneous,
                                  downreg_list$DefScl,
                                  downreg_list$DefScl_NonCutaneous,
                                  downreg_list$DefScl_Limited,
                                  downreg_list$DefScl_Diffuse)),
                         Sjogren = 
                         downreg_list$Sjogren,
                         Arthritis = 
                         downreg_list$RheumArthritis,
                         Dermatomyositis = 
                         downreg_list$JuvDermatomyositis
)

#### Overlaps

# Upregulated DEGs

upset(fromList(merged_upreg_list), 
      order.by = "freq",
      nsets = 5,
      main.bar.color = "steelblue", 
      sets.bar.color = "tomato")

# Downregulated DEGs

upset(fromList(merged_downreg_list), 
      order.by = "freq",
      nsets = 5,
      main.bar.color = "steelblue", 
      sets.bar.color = "tomato")

# Comparative summary

lapply(merged_upreg_list, length)

length(unique(c(merged_upreg_list$SLE,
                merged_upreg_list$Sclerosis,
                merged_upreg_list$Sjogren,
                merged_upreg_list$Arthritis,
                merged_upreg_list$Dermatomyositis)))

length(c(merged_upreg_list$SLE,
         merged_upreg_list$Sclerosis,
         merged_upreg_list$Sjogren,
         merged_upreg_list$Arthritis,
         merged_upreg_list$Dermatomyositis))

lapply(merged_downreg_list, length)

length(unique(c(merged_downreg_list$SLE,
                merged_downreg_list$Sclerosis,
                merged_downreg_list$Sjogren,
                merged_downreg_list$Arthritis,
                merged_downreg_list$Dermatomyositis)))

length(c(merged_downreg_list$SLE,
         merged_downreg_list$Sclerosis,
         merged_downreg_list$Sjogren,
         merged_downreg_list$Arthritis,
         merged_downreg_list$Dermatomyositis))
```

### Subtype Overlap

While the global disease overlap tells us what is shared across different autoimmune conditions, it is interesting to address the internal heterogeneity of specific diseases at a cellular level. For instance, the GSE249641 dataset for SLE separates monocytes into three distinct functional subsets: Classical, Intermediate, and Non-Classical. By intersecting the DEGs of these specific subsets, we can determine whether the SLE transcriptomic signature is a pan-monocyte phenomenon or driven predominantly by a single differentiation state.

For this section, the FC threshold is set at the standard 2x. The increased resolution of this approach leaded to an artificially enlarged DEG list, while a lower threshold was useful in the rest of the studies to retrieve more subtle regulatory patterns.

```{r}
# Subtypes lists
subtype_upreg = list(Classical = upreg_list$SLE_Classical,
  Intermediate = upreg_list$SLE_Intermediate,
  NonClassical = upreg_list$SLE_NonClassical)
  
subtype_downreg = list(Classical = downreg_list$SLE_Classical,
  Intermediate = downreg_list$SLE_Intermediate,
  NonClassical = downreg_list$SLE_NonClassical)

# Overlap

ggVennDiagram(subtype_upreg, 
              label_alpha = 0,
              label_color = "black",
              edge_lty = "solid",
              edge_size = 0.8
              ) + 
  scale_fill_distiller(palette = "YlOrRd", direction = 1
                       ) + 
  theme(
  legend.position = "bottom",
  legend.title = element_text(face = "bold"),
  legend.key.width = unit(1.5, "cm"),
       ) +
  labs(title = "Upregulated DEGs by Monocyte Subtype",
       fill = "DEGs")

ggVennDiagram(subtype_downreg, 
              label_alpha = 0,
              label_color = "black",
              edge_lty = "solid",
              edge_size = 0.8
              ) + 
  scale_fill_distiller(palette = "Blues", direction = 1
                       ) + 
  theme(
  legend.position = "bottom",
  legend.title = element_text(face = "bold"),
  legend.key.width = unit(1.5, "cm"),
       ) +
  labs(title = "Downregulated DEGs by Monocyte Subtype",
       fill = "DEGs")

###### CHANGE THRESHOLDS

# Subtypes lists
subtype_upreg = list(Classical = extract_DEGs(SLE_GSE249641$classical$res, "up", lfc_th = 1),
  Intermediate = extract_DEGs(SLE_GSE249641$intermediate$res, "up", lfc_th = 1),
  NonClassical = extract_DEGs(SLE_GSE249641$non_classical$res, "up", lfc_th = 1))
  
subtype_downreg = list(Classical = extract_DEGs(SLE_GSE249641$classical$res, "down", lfc_th = 1),
  Intermediate = extract_DEGs(SLE_GSE249641$intermediate$res, "down", lfc_th = 1),
  NonClassical = extract_DEGs(SLE_GSE249641$non_classical$res, "down", lfc_th = 1))

# Overlap

ggVennDiagram(subtype_upreg, 
              label_alpha = 0,
              label_color = "black",
              edge_lty = "solid",
              edge_size = 0.8
              ) + 
  scale_fill_distiller(palette = "YlOrRd", direction = 1
                       ) + 
  theme(
  legend.position = "bottom",
  legend.title = element_text(face = "bold"),
  legend.key.width = unit(1.5, "cm"),
       ) +
  labs(title = "Upregulated DEGs by Monocyte Subtype",
       fill = "DEGs")

ggVennDiagram(subtype_downreg, 
              label_alpha = 0,
              label_color = "black",
              edge_lty = "solid",
              edge_size = 0.8
              ) + 
  scale_fill_distiller(palette = "Blues", direction = 1
                       ) + 
  theme(
  legend.position = "bottom",
  legend.title = element_text(face = "bold"),
  legend.key.width = unit(1.5, "cm"),
       ) +
  labs(title = "Downregulated DEGs by Monocyte Subtype",
       fill = "DEGs")
```

### MetaScape Visualization

Web tools like MetaScape are often superior for generating highly stylized, publication-ready functional enrichment visualizations. In this section, we mainly focus on "functional clustering" tools, based on shared DEGs and represented as heatmaps. MetaScape requires simple CSV file where each column is a gene list.

```{r}
# Fill NAs for equal sizes
upreg_metascape = data.frame(lapply(merged_upreg_list, 
                             function(x) {
      length(x) = max_len = max(sapply(merged_upreg_list, length)) 
      x[is.na(x)] = ""    
      return(x)
      }), 
      check.names = FALSE)

downreg_metascape = data.frame(lapply(merged_downreg_list, 
                             function(x) {
      length(x) = max_len = max(sapply(merged_downreg_list, length)) 
      x[is.na(x)] = ""    
      return(x)
      }), 
      check.names = FALSE)

up_sub_metascape = data.frame(lapply(subtype_upreg, 
                             function(x) {
      length(x) = max_len = max(sapply(subtype_upreg, length)) 
      x[is.na(x)] = ""    
      return(x)
      }), 
      check.names = FALSE)

down_sub_metascape = data.frame(lapply(subtype_downreg, 
                             function(x) {
      length(x) = max_len = max(sapply(subtype_downreg, length)) 
      x[is.na(x)] = ""    
      return(x)
      }), 
      check.names = FALSE)

## Export the data frames
write.csv(upreg_metascape, 
          file = "data/Outputs/Metascape_Upreg_Input.csv", 
          row.names = FALSE, 
          quote = FALSE)

write.csv(downreg_metascape, 
          file = "data/Outputs/Metascape_Downreg_Input.csv", 
          row.names = FALSE, 
          quote = FALSE)

write.csv(up_sub_metascape, 
          file = "data/Outputs/Metascape_UpregSubtype_Input.csv", 
          row.names = FALSE, 
          quote = FALSE)

write.csv(down_sub_metascape, 
          file = "data/Outputs/Metascape_DownregSubtype_Input.csv", 
          row.names = FALSE, 
          quote = FALSE)
```

# Meta-Analysis (Late Integration)

While simple overlaps are visually intuitive, they are statistically fragile. For example, if a crucial gene strongly passes the significance threshold in 4 diseases, but narrowly misses it in the 5th due to a smaller sample size, a strict Venn diagram will completely discard that gene from the "Core Signature".

To overcome this, we perform a formal meta-analysis using MetaVolcanoR. The algorithm uses a Random Effects Model, which calculates a weighted average of the expression changes, giving more weight to highly precise cohorts and penalizing noisy ones.

### Data Preparation

MetaVolcanoR requires a highly specific format, including genes, logFC, p-values and 95% confidence intervals (lower and upper limits). Notably, DESeq2 provides the standard error of the log2FC (lfcSE) but the meta-analysis requires the confidence intervals, which can be calculated from it.

```{r}
library(MetaVolcanoR)

# DESeq2 results were saved as df
# Columns: baseMean, log2FC, lfcSE, pvalue and padj
# Rownames: ENSEMBL annotation

# Prepare results function
prepare_res = function(res_df) {
  
  # Necessary columns with their standardized names for MetaVolcanoR
  named_df = data.frame(
    Gene    = rownames(res_df),
    logFC   = res_df$log2FoldChange,
    P.Value = res_df$pvalue,
    CI.L    = res_df$log2FoldChange - (1.96 * res_df$lfcSE),
    CI.R    = res_df$log2FoldChange + (1.96 * res_df$lfcSE)
  )
  
  # Clean NAs to prevent error
  clean_df = na.omit(named_df)
  return(clean_df)
}
```

### Intra-disease Pre-Meta-Analysis

Before we can compare the diseases, we must collapse the multi-cohort diseases (SLE and Scleroderma) into single, unified profiles to avoid overrepresentation of these diseases. We run a "Pre-Meta-Analysis" to fuse the subtypes together, obtaining meta-statistics that will then be used as the original ones.

Note: There is a known bug in R environments regarding S4 object validation when using MetaVolcanoR, due to inconsistencies between R, ggplot2 and MetaVolcanoR versions. We bypassed this by overriding the ValidObject namespace, but compilation cannot be fulfilled directly when this action is executed during the process (saved as RData and eval=F for compilation).

```{r, eval=F}
#### SLE
 # For GSE149050, SLE_IFNneg is discarded due to lack of significant differences against hc
 # For GSE249641, we integrate the subtypes as global design added noise
list_SLE = list(
  SLE_IFNpos = prepare_res(SLE_GSE149050$res_SLE_IFNpos),
  SLE_Classical = prepare_res(SLE_GSE249641$classical$res),
  SLE_Intermediate = prepare_res(SLE_GSE249641$intermediate$res),
  SLE_Nonclassical = prepare_res(SLE_GSE249641$non_classical$res)
)

# Added to solve versions incompatibilities 
library(methods)
setValidity("MetaVolcano", function(object) TRUE)
unlockBinding("validObject", as.environment("package:methods"))
assignInNamespace("validObject",
                  function(object, 
                           test = FALSE, complete = FALSE) TRUE,
                  ns = "methods")

# Random Effects Meta-Analysis
meta_rem_SLE = rem_mv(diffexp      = list_SLE,
                      pcriteria    = "P.Value",
                      foldchangecol= "logFC",
                      genenamecol  = "Gene",
                      collaps      = TRUE,
                      llcol        = "CI.L",
                      rlcol        = "CI.R",
                      metathr      = 0.01,
                      jobname      = "SLE_MetaVolcano",
                      outputfolder = "data/Outputs/",
                      draw         = "PDF")

save(meta_rem_SLE,
     file = "data/Outputs/SLE_MetaVolcano_res.RData")


#### Scl
 # GSE303431 is discarded, as no DEGs were found at all
 # For GSE124073 we take the independent cohorts, discarding the early stage as it is indistinguishable from hc
list_Scl = list(
  NFScl_NonCutaneous = prepare_res(Scl_GSE124073$Milan$subtypes$res_noncut),
  DefScl_NonCutaneous = prepare_res(Scl_GSE124073$Utrecht$subtypes$res_noncut),
  DefScl_Limited = prepare_res(Scl_GSE124073$Utrecht$subtypes$res_limited),
  DefScl_Diffuse = prepare_res(Scl_GSE124073$Utrecht$subtypes$res_diffuse)
)

# Added to solve versions incompatibilities 
library(methods)
setValidity("MetaVolcano", function(object) TRUE)
unlockBinding("validObject", as.environment("package:methods"))
assignInNamespace("validObject",
                  function(object, 
                           test = FALSE, complete = FALSE) TRUE,
                  ns = "methods")

meta_rem_Scl = rem_mv(diffexp      = list_Scl,
                      pcriteria    = "P.Value",
                      foldchangecol= "logFC",
                      genenamecol  = "Gene",
                      collaps      = TRUE,
                      llcol        = "CI.L",
                      rlcol        = "CI.R",
                      metathr      = 0.01,
                      jobname      = "Scl_MetaVolcano",
                      outputfolder = "data/Outputs/",
                      draw         = "PDF")

save(meta_rem_Scl,
     file = "data/Outputs/Scl_MetaVolcano_res.RData")
```

```{r}
# Visualize results
load("data/Outputs/SLE_MetaVolcano_res.RData")
meta_rem_SLE@MetaVolcano

load("data/Outputs/Scl_MetaVolcano_res.RData")
meta_rem_Scl@MetaVolcano
```

### Meta-Analysis: Core integration

With SLE and Scleroderma safely collapsed into singular, robust meta-profiles, we can now integrate all 5 distinct autoimmune diseases. First, we need to extract the newly generated consensus logFCs and p-values from our SLE and Scleroderma "pre-meta-analyses" so they match the format of the single-cohort diseases (Sjögren, Arthritis, Dermatomyositis).

Finally, we run the ultimate combinatorial meta-analysis. This uses a combination approach (like Fisher's method) to derive a single, overarching metric for every gene across all 5 conditions.

```{r}
# Adapt format as single study for SLE and Scl
# 'randomSummary' = Combined log2FC and 'randomP' = Combined p-value
meta_SLE_def = meta_rem_SLE@metaresult

meta_SLE_def = data.frame(
  Gene    = meta_SLE_def$Gene,
  logFC   = meta_SLE_def$randomSummary,
  P.Value = meta_SLE_def$randomP,
  CI.L    = meta_SLE_def$randomCi.lb,
  CI.R    = meta_SLE_def$randomCi.ub
)

meta_Scl_def = meta_rem_Scl@metaresult

meta_Scl_def = data.frame(
  Gene    = meta_Scl_def$Gene,
  logFC   = meta_Scl_def$randomSummary,
  P.Value = meta_Scl_def$randomP,
  CI.L    = meta_Scl_def$randomCi.lb,
  CI.R    = meta_Scl_def$randomCi.ub
)

## Definitive list

list_all = list(
  SLE = meta_SLE_def,
  Sclerosis = meta_Scl_def,
  Sjogren = prepare_res(SS_GSE173670$res_pSS),
  Arthritis = prepare_res(RA_GSE294225$res),
  Dermatomyositis = prepare_res(JD_GSE221091$res)
)
```

```{r, eval=F}
# Added to solve versions incompatibilities 
library(methods)
setValidity("MetaVolcano", function(object) TRUE)
unlockBinding("validObject", as.environment("package:methods"))
assignInNamespace("validObject",
                  function(object, 
                           test = FALSE, complete = FALSE) TRUE,
                  ns = "methods")

# Combinatory!
meta_global_ALL = combining_mv(diffexp       = list_all,
                               pcriteria     = "P.Value",
                               foldchangecol = "logFC",
                               genenamecol   = "Gene",
                               metafc        = "Mean",
                               collaps       = TRUE,
                               metathr       = 0.01,
                               jobname       = "ALL_MetaVolcano",
                               outputfolder  = "data/Outputs/",
                               draw          = "PDF")

save(meta_global_ALL,
     file = "data/Outputs/ALL_MetaVolcano_res.RData")
```

```{r}
load("data/Outputs/ALL_MetaVolcano_res.RData")
meta_global_ALL@MetaVolcano

### MANUAL METAVOLCANO

library(tidyr)
library(dplyr)

# Classify
mr_filtered = meta_global_ALL@metaresult |>
  filter(!is.na(metafc) & !is.na(metap)) |>
  mutate(
    neg_log10_p = -log10(metap),
    diff_state = case_when(
      metafc > log2(1.5)  & metap < 0.01 ~ "Activated",
      metafc < -log2(1.5)  & metap < 0.01 ~ "Repressed",
      TRUE ~ "Not significant"
    ),
    diff_state = factor(diff_state, levels = c("Repressed", "Activated", "Not significant"))
  )

n_activated = sum(mr_filtered$diff_state == "Activated", na.rm = TRUE)
n_repressed = sum(mr_filtered$diff_state == "Repressed", na.rm = TRUE)

# Plotting
ggplot(mr_filtered, aes(x = metafc, y = neg_log10_p, color = diff_state)) +
  geom_point(alpha = 0.5, size = 1.2) +
  scale_color_manual(values = c("Activated"       = "red",
                                "Repressed"       = "blue",
                                "Not significant" = "grey70")) +
  geom_vline(xintercept = c(-log2(1.5), log2(1.5)),
             linetype = "dashed", color = "black", alpha = 0.5) +
  geom_hline(yintercept = -log10(0.01),
             linetype = "dashed", color = "black", alpha = 0.5) +
  labs(title    = "Meta-Volcano - Core Signature",
       subtitle = paste("Repressed", n_repressed, "| Activated:", n_activated),
       x        = "log2 Fold Change",
       y        = "-log10 metap",
       color    = "") +
  theme_minimal(base_size = 14) +
  theme(legend.position  = "bottom",
        panel.grid.minor = element_blank())
```

Mean combinatorial meta-volcano shows abundant noise. Using a weighted meta-analysis approach offers a significant statistical advantage because it intrinsically accounts for the reliability and precision of each individual study. While a simple mean approach treats every dataset equally, regardless of its sample size or internal variance, a weighted method applies inverse variance weighting, granting more influence to high-quality studies while penalizing noisy ones. This critical adjustment prevents a single low-powered experiment or extreme outlier from disproportionately skewing the consensus results. Ultimately, prioritizing study precision yields a much more robust and statistically powerful summary estimate, substantially reducing false positives and improving the reliability of the molecular signatures across heterogeneous datasets.

```{r, eval=F}
# Precision-weighted metafc, based on inverse variance
# SE is recovered from the 95% CI
# Weight = 1 /SE^2
# Weighted FC = sum(weight * FC) / sum(weight) per gene across diseases

# Get intermediate values
mr = meta_global_ALL@metaresult

diseases = c("SLE", "Sclerosis", "Sjogren", "Arthritis", "Dermatomyositis")

compute_weighted_fc = function(gene, list_all, diseases) {
  fcs = sapply(diseases, function(d) {
    row = list_all[[d]][list_all[[d]]$Gene == gene, ]
    if (nrow(row) == 0) return(NA_real_)
    row$logFC[1]
  })
  ses = sapply(diseases, function(d) {
    row = list_all[[d]][list_all[[d]]$Gene == gene, ]
    if (nrow(row) == 0) return(NA_real_)
    ((row$CI.R[1] - row$CI.L[1]) / (2 * 1.96))
  })
  w = 1 / ses^2
  w[!is.finite(w)] = NA
  if (all(is.na(w))) return(NA_real_)
  sum(w * fcs, na.rm = TRUE) / sum(w, na.rm = TRUE)
}

# Compute final weighted FC
weighted_fc_vec = sapply(mr$Gene, compute_weighted_fc, 
                         list_all = list_all, 
                         diseases = diseases)

# Avoid corrupt data. Minimum of 3 diseases covered
coverage = sapply(mr$Gene, function(gene) {
  sum(sapply(diseases, function(d) {
    row = list_all[[d]][list_all[[d]]$Gene == gene, ]
    nrow(row) > 0 && is.finite(row$CI.R[1]) && is.finite(row$CI.L[1])
  }))
})

weighted_fc_vec[coverage < 3] = NA


# Replace naive mean with weighted estimate
meta_global_ALLweighted = meta_global_ALL
meta_global_ALLweighted@metaresult$metafc = weighted_fc_vec
meta_global_ALLweighted@metaresult$coverage = coverage

# Save results
save(meta_global_ALLweighted,
     file = "data/Outputs/ALLweighted_MetaVolcano_res.RData")
```

### Extract results - Metascape visualization

We extract the final dataframe, which contains a metap (the combined global p-value) and a metafc (the combined fold change). We apply our strict significance filters here one last time, resulting in our definitive Autoimmunity Core Signature.

Finally, we translate the Ensembl IDs to standard HUGO Gene Symbols so the list can be exported as a CSV and immediately interpreted by scientists or fed into functional pathway tools. Especially, we will focus on Metascape functional analysis.

```{r}
load("data/Outputs/ALLweighted_MetaVolcano_res.RData")  

# See final dataframe 
final_res = meta_global_ALLweighted@metaresult 
head(final_res)  

# Integrated p-value and mean LogFC, removing low covered
filtered_res = subset(final_res, 
                      metap < 0.01 & 
                      metafc > log2(1.5) &
                      !is.na(metafc))

# Add gene symbols translation 
library(org.Hs.eg.db)  
filtered_res$Symbol = mapIds(x = org.Hs.eg.db,   
                             keys = filtered_res$Gene,   
                             column = "SYMBOL",   
                             keytype = "ENSEMBL",   
                             multiVals = "first" )  

# Clean NAs 
export_res = filtered_res[!is.na(filtered_res$Symbol), ]  
dim(export_res)

# Export to csv for MetaScape 
write.csv(export_res, file = "data/Outputs/AutoimmunityCore_DEGs.csv", row.names = FALSE)
```

### Additional results

To validate the precision-weighting strategy, some additional graphs can be interesting:

-   Average weight per disease. The CI size and the weight given are inversely proportional. We can check the per-disease "strength".

-   Naive mean vs weighted FC. This plot gives us a visual explanation on why weighting the FC is necessary.

-   Behavior for top-10 metaDEGs.

-   Meta-volcano plot for the new approach.

```{r}
library(tidyr)
library(dplyr)

# Consistent colors
diseases = c("SLE", "Sclerosis", "Sjogren", "Arthritis", "Dermatomyositis")

disease_colors = c(
  "SLE"              = "chartreuse4",
  "Sclerosis"        = "darkorchid3",
  "Sjogren"          = "darkorange1",
  "Arthritis"        = "firebrick2",  
  "Dermatomyositis"  = "deepskyblue2",
  "Weighted\nSummary"= "black"
)


##### AVERAGE WEIGHT PER DISEASE

# Calculate intermediate values
mean_w_df = lapply(diseases, function(d) {
                df  = list_all[[d]]
                se  = (df$CI.R - df$CI.L) / (2 * 1.96)
                w   = 1 / se^2
                w[!is.finite(w)] = NA
                data.frame(Disease = d, MeanWeight = mean(w, na.rm = TRUE))
              }) |> bind_rows()

# Plotting
ggplot(mean_w_df, 
       aes(x = reorder(Disease, MeanWeight), 
           y = MeanWeight, 
           fill = Disease)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = round(MeanWeight, 2)), 
            hjust = -0.15, 
            size = 3.8) +
  coord_flip() +
  scale_fill_manual(values = disease_colors) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title    = "Average Inverse-Variance Weight by Disease",
       x = NULL, 
       y = "Mean weight (1 / SE²)") +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank())


##### NAIVE MEAN VS WEIGTHED FC SCATTER
mr = meta_global_ALLweighted@metaresult

# Calculate mean by gene
all_df = bind_rows(list_all[diseases], .id = "Disease")

naive_df = all_df |>
  group_by(Gene) |>
  summarize(naive = mean(logFC, na.rm = TRUE), .groups = "drop")

# Mean vs meta-analysis results
comparison_df = naive_df |>
  inner_join(mr[, c("Gene", "metafc")], by = "Gene") |>
  rename(weighted = "metafc") |>
  filter(is.finite(naive) & is.finite(weighted))

# R-squared 
r2 = round(cor(comparison_df$naive, 
               comparison_df$weighted, 
               use = "complete.obs")^2, 3)

# Plotting
ggplot(comparison_df, 
       aes(x = naive, 
           y = weighted,
           color = weighted > log2(1.5))) +
  geom_point(alpha = 0.75, size = 1.5) +
  geom_abline(slope = 1, intercept = 0, color = "firebrick", linewidth = 0.6, alpha = 0.6) +
  geom_hline(yintercept = log2(1.5),
             linetype = "dashed", color = "grey20", linewidth = 0.6, alpha = 0.8) +
  geom_vline(xintercept = log2(1.5),
             linetype = "dashed", color = "grey20", linewidth = 0.6, alpha = 0.8) +
  annotate("text", x = -Inf, y = Inf, hjust = -4.5, vjust = 5.5,
           label = paste0("R² = ", r2), size = 4.5, col = "firebrick") +
  scale_color_manual(values = c("TRUE" = "coral", "FALSE" = "grey60")) +
  labs(title = "Naive Mean vs Precision-Weighted logFC",
       x = "Naive mean logFC", 
       y = "Weighted logFC") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")


##### FOREST PLOT

# Top-10 genes (by idx)
top_genes = export_res |>
  arrange(desc(idx)) |>
  head(10) |>
  select(Gene, Symbol, metafc, metap, idx)

# Per-disease information
forest_df = all_df |>
  filter(Gene %in% top_genes$Gene) |>
  select(Disease, Gene, logFC, CI.L, CI.R) |>
  left_join(top_genes[, c("Gene", "Symbol", "metafc")], by = "Gene") |>
  mutate(Disease = factor(Disease, levels = diseases))

# Meta-summary information
summary_df = top_genes |>
  mutate(Disease = "Weighted\nSummary",
         logFC   = metafc,
         CI.L    = NA_real_,
         CI.R    = NA_real_) |>
  select(Gene, Symbol, logFC, CI.L, CI.R, Disease, metafc) 

# Join and plotting
forest_full = bind_rows(forest_df, summary_df) |>
  mutate(
    Disease = factor(Disease, levels = c(diseases, "Weighted\nSummary")),
    IsSum   = Disease == "Weighted\nSummary",
    Symbol  = factor(Symbol, levels = rev(top_genes$Symbol)),
    Panel   = factor(ifelse(IsSum, "Weighted Summary", "Per-disease logFC"),
                     levels = c("Per-disease logFC", "Weighted Summary"))
  )

dummy_limits = data.frame(
  Panel = rep(levels(forest_full$Panel), each = 2),
  logFC = c(-1, 9, 0, 3.5),
  Symbol = forest_full$Symbol[1]
)

ggplot(forest_full,
       aes(x = logFC, 
           y = Symbol, 
           color = Disease,
           shape = IsSum,
           size  = IsSum)) +
  
  # Size and grey threshold
  geom_blank(data = dummy_limits, 
             aes(x = logFC, y = Symbol), inherit.aes = FALSE) +
  scale_x_continuous(breaks = seq(-1, 9, by = 1), expand = c(0, 0)) + 
  facet_grid(. ~ Panel, scales = "free_x", space = "free_x") +
  annotate("rect", 
           xmin = -Inf, xmax = log2(1.5), 
           ymin = -Inf, ymax = Inf, 
           fill = "grey80", alpha = 0.4) +
  geom_vline(xintercept = log2(1.5), 
             linetype = "solid", color = "grey60", linewidth = 0.4) +

  # Values
  geom_errorbarh(
    aes(xmin = CI.L, xmax = CI.R),
    height = 0.2, linewidth = 0.6,
    data = filter(forest_full, !IsSum)
  ) +
  geom_point() +
  scale_shape_manual(values = c("FALSE" = 16, "TRUE"  = 18)) +
  scale_size_manual(values = c("FALSE" = 1.8,"TRUE"  = 4.2)) +
  scale_color_manual(values = disease_colors) +

  # Format
  labs(
    title = "Top-10 Core Signature Genes (metap x metaFC)",
    x = "log2 Fold Change", 
    y = NULL,
    color = "Disease"
  ) +
  guides(
    shape = "none", 
    size  = "none", 
    color = guide_legend(nrow = 1)
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position   = "bottom",
    legend.title      = element_text(face = "bold", size = 11),
    legend.text       = element_text(size = 9),
    strip.text        = element_text(face = "bold", size = 12.5),
    panel.grid.minor  = element_blank(),
    panel.grid.major.y= element_line(linetype = "dotted", color = "grey88"),
    panel.border      = element_rect(color = "black", fill = NA, linewidth = 0.7),
    axis.title.x      = element_text(size = 12),
    panel.spacing     = unit(1, "cm")
  )

##### MANUAL METAVOLCANO BY IDX

# Classify 
mr_filtered = meta_global_ALLweighted@metaresult |>
  filter(!is.na(metafc) & !is.na(metap)) |>
  mutate(
    neg_log10_p = -log10(metap),
    diff_state = case_when(
      metafc > log2(1.5)  & metap < 0.01 ~ "Activated",
      TRUE ~ "Not significant"
    ),
    diff_state = factor(diff_state, levels = c("Activated", "Not significant"))
  )

n_activated = sum(mr_filtered$diff_state == "Activated", na.rm = TRUE)

# Get labels por top-10 by idx
mr_filtered = mr_filtered |>
  mutate(
    ensembl_label = ifelse(Gene %in% top_genes$Gene, Gene, NA),
    label = mapIds(org.Hs.eg.db,
                   keys = ensembl_label,
                   column = "SYMBOL",
                   keytype = "ENSEMBL",
                   multiVals = "first")
  )

# Plotting
ggplot(mr_filtered, aes(x = metafc, y = neg_log10_p, color = diff_state)) +
  geom_point(alpha = 0.5, size = 1.2) +
  scale_color_manual(values = c("Activated"       = "red",
                                "Not significant" = "grey70")) +
  geom_vline(xintercept = c(-log2(1.5), log2(1.5)), 
             linetype = "dashed", color = "black", alpha = 0.5) +
  geom_hline(yintercept = -log10(0.01), 
             linetype = "dashed", color = "black", alpha = 0.5) +
  ggrepel::geom_label_repel(aes(label = label),
                            size = 3.5, max.overlaps = 30,
                            na.rm = TRUE, color = "black",
                            fill = alpha("white", 0.7)) +
  labs(title    = "Weighted Meta-Volcano - Core Signature",
       subtitle = paste("Activated:", n_activated),
       x        = "Weighted log2 Fold Change", 
       y        = "-log10 metap",
       color    = "") +
  theme_minimal(base_size = 14) +
  theme(legend.position  = "bottom",
        panel.grid.minor = element_blank())
```

#### Session Info

```{r}
sessionInfo()
```
